Uncharacterized yeast gene YBR238C, an effector of TORC1 signaling in a mitochondrial feedback loop, accelerates cellular aging via HAP4- and RMD9-dependent mechanisms

Version 1

Abstract

Summary

Uncovering the regulators of cellular aging will unravel the complexity of aging biology and identify potential therapeutic interventions to delay the onset and progress of chronic, aging-related diseases. In this work, we systematically compared gene sets involved in regulating the lifespan of Saccharomyces cerevisiae (a powerful model organism to study the cellular aging of humans) and those with expression changes under rapamycin treatment. Among the functionally uncharacterized genes in the overlap set, YBR238C stood out as the only one downregulated by rapamycin and with an increased chronological and replicative lifespan upon deletion. We show that YBR238C and its paralogue RMD9 oppositely affect mitochondria and aging. YBR238C deletion increases the cellular lifespan by enhancing mitochondrial function. Its overexpression accelerates cellular aging via mitochondrial dysfunction. We find that the phenotypic effect of YBR238C is largely explained by HAP4– and RMD9-dependent mechanisms. Further, we find that genetic or chemical-based induction of mitochondrial dysfunction increases TORC1 (Target of Rapamycin Complex 1) activity that, subsequently, accelerates cellular aging. Notably, TORC1 inhibition by rapamycin (or deletion of YBR238C) improves the shortened lifespan under these mitochondrial dysfunction conditions in yeast and human cells. The growth of mutant cells (a proxy of TORC1 activity) with enhanced mitochondrial function is sensitive to rapamycin whereas the growth of defective mitochondrial mutants is largely resistant to rapamycin compared to wild type. Our findings demonstrate a feedback loop between TORC1 and mitochondria (the TORC1-MItochondria-TORC1 (TOMITO) signaling process) that regulates cellular aging processes. Hereby, YBR238C is an effector of TORC1 modulating mitochondrial function.

Introduction

Healthy aging is crucially determined by cellular functions, and their defective status is associated with premature dysfunction and/or depletion of critical cell populations and various aging-associated pathologies such as neurodegenerative diseases, cancer, cardiovascular disorders, diabetes, sarcopenia and maculopathy 1–6. Advancements in biomedical research including genome-wide screening in different aging model organisms, identified several biological pathways that support the unprecedented progress in understanding overlapping profiles between aged cells and different chronic aging-associated diseases 7–10. In principle, uncovering the molecular mechanisms that drive cellular aging identifies potential drug targets and can fuel the development of therapeutics to delay aging and increase healthspan 11–14.

Given the evolutionary conservation of many aging-related pathways, yeast is one of the aging model organisms that have been extensively used to study the biology of human cellular aging by analyzing chronological lifespan (CLS) and replicative lifespan (RLS) under various conditions 15–18. The CLS is the duration of time that a non-dividing cell is viable. This is a cellular aging model for post-mitotic human cells such as neurons and muscle cells. The RLS defines as the number of times a mother cell divides to form daughter cells. Such experiments provide a replicative human aging model for mitotic cells such as stem cells. Genome-wide or individual gene deletion strains’ screening identified thousands of genes affecting cellular lifespan. These lists are a rich resource to identify unique genetic regulators, functional networks, and interactions of aging hallmarks relevant to cellular lifespan 1,4,7,11–14. At the same, these lists are rich in genes of unknown function, a class of genes that, unfortunately, got increasingly ignored by the attention of research teams 19–21.

We systematically examined the available genetic data on aging and lifespan for budding yeast, Saccharomyces cerevisiae, from various sources. Our bioinformatics analyses revealed consensus lists of genes regulating yeast CLS and RLS. These gene sets were compared with lists of genes the expression of which changed under TORC1 inhibition with rapamycin. When we turned our attention to the functionally uncharacterized genes involved, we found YBR238C the only one among the latter that increases both CLS and RLS upon deletion and that is downregulated by rapamycin. Transcriptomics and biochemical experiments revealed that YBR238C negatively regulates mitochondrial function, largely via HAP4– and RMD9– dependent mechanisms, and thereby affects cellular lifespan. Surprisingly, YBR238C and its paralogue RMD9 oppositely influence mitochondrial function and cellular aging.

Our chemical genetics and metabolic analyses unravel a feedback loop of the interaction of TORC1 with mitochondria that affect cellular aging. YBR238C is an effector of TORC1 that modulates mitochondrial function. We also show that mitochondrial dysfunction induces TORC1 activity enhancing cellular aging. In turn, TORC1 inhibition in yeast and human cells with mitochondrial dysfunction suppresses their accelerated cellular aging.

Results

Genome-wide survey of genes affecting cellular lifespan in yeast in accordance with literature and public databases

Lists of scientific literature instances mentioning a yeast gene as affecting lifespan were downloaded from the databases SGD (Saccharomyces cerevisiae genome database) 22, and GenAge 23. In most cases, the experiments referred to are gene deletion phenotype studies. After processing the files for the mentioning of ‘increase/decrease’ of ‘chronological lifespan’ (CLS) and/or ‘replicative lifespan’ (RLS) as well as the suppression of gene duplicates, we found 2399 entries with distinct yeast genes in 15 categories reported to increase/decrease CLS and/or RLS under various conditions (Table 1). We collectively call them Aging Associated Genes (AAGs). Notably, about one-third of the total yeast genome belongs to that category. Downloaded files (as of 8th November 2022), description of the processing details, and all the resulting gene lists are available in ‘Additional File 1’.

Whereas most of the genes (1610, 67%) have been mentioned for just one of the four scenarios (“CLS increase”, “CLS decrease”, “RLS increase”, and “RLS decrease”), the remaining genes have been described for several alternative and/or even opposite/conflicting outcomes. 19 genes are brought up in context with all four scenarios. As the experimental conditions varied among the reports and the gene networks are complex, this is not necessarily a contradiction. Yet (see below) some of the cases are actually annotation errors at the database level.

We explored the potential enrichment of gene ontology (GO) terms among the genes involved in the 15 categories with the help of DAVID WWW server 24. The most significant result is the enrichment of ontology terms related to mitochondrial function among the genes annotated with the single qualifier “CLS decrease”. The top mitochondrial term cluster has an enrichment score 18.2 (gene-wise P-values and Benjamini-Hochberg values all below 6.e-13); a second one related to mitochondrial translation has the score 4.7. Enrichment of mitochondrial ontology terms has also been observed for “CLS decrease” in combination with “CLS increase” or “RLS increase/decrease”.

Every other signal from the ontology is much weaker. Term enrichment (with scores near 2 or better) related to autophagy and vesicular transport pop up for CLS-annotated genes whereas terms connected with translation, DNA repairs, telomeres, protein degradation and signaling are observed with RLS-tagged gene lists.

We also explored the presence of uncharacterized or severely under-characterized genes in the 15 categories of AAGs (Table 1). In total, 944 genes are annotated in SGD as coding for a protein of unknown function. Additionally, we considered genes without dedicated gene name (only with six-letter locus tag) as dramatically under-characterized. Comparison of this combined list with those in the 15 categories reveals 88 severely understudied AAGs that are candidates for enhanced attention from the scientific community. Thus, functionally insufficiently characterized genes have a great role in cellular aging-related processes.

Rapamycin response genes overlap with the AAGs

Nutrient sensing dysregulation is one of the aging hallmarks 11,12. The conserved protein complex Target of Rapamycin Complex 1 (TORC1) senses nutrients such as amino acids and glucose and links metabolism with cellular growth and proliferation 25–28. TORC1 positively regulates aging, and its inhibition increases lifespan in various eukaryotic organisms including yeast and mammalian 13,26,27,29,30. The drug rapamycin, initially discovered as an antifungal natural product produced by Streptomyces hygroscopicus, inhibits TORC1 and increases lifespan (Figures S1A and S1B) 13,17,19,21,23,25,29.

To explore the connection between nutrient signaling and cellular aging we mapped the TORC1 regulated genes with AAGs. We first identified rapamycin response genes (RRGs) by transcriptomics analysis (RNA-Seq for yeast S. cerevisiae BY4743 cells treated with rapamycin and DMSO control; Figure S1C). Relevant measurement results, lists of 2365 RRGs and supplementary methodical comments are available in ‘Additional File 2’.

As overlap of two RNA-Seq data analysis methods (see Methods), we identified 1155 rapamycin upregulate genes (RUG) and 1210 rapamycin downregulated genes (RDG) (Figures S1D-S1F; Additional File 2). RNA-Seq results were confirmed for some genes by qRT-PCR (Figure S2A). We also checked the differential gene expressions in prototrophic yeast S. cerevisiae strain CEN.PK and found similar results as with the BY4743 strain (Figure S2B). To note, our transcriptomics analyses are, as a trend, consistent with previous RNA-Seq studies carried out under partially different experimental conditions 37,38.

We mapped the RRGs (RUG and RDG) with AAGs. Among the 2399 AAGs names, 397 and 433 (in total 830) re-occur in the lists of RUG and RDG, respectively. Thus, the overlap with AAGs is nearly 35%. I.e., TORC1 controls about one third of the AAGs. Table 1 shows how many AAGs are up– and down-regulated by rapamycin treatment for each of the 15 categories with regard to CLS/RLS increase/decrease. Notably, the order of magnitude for the respective numbers of RUG and RDG is the same for all 15 studied subgroups.

Since rapamycin increases the lifespan by an inhibitory effect on TORC1 activity 13,17,19,21,23,25,29, it appears most interesting to focus on AAGs with a deletion phenotype of increased CLS/RLS and being down-regulated by rapamycin application. A manual analysis of these gene lists reveals that, among the uncharacterized or severely understudied AAGs, there is a single one known to increase both CLS and RLS upon deletion and being downregulated by rapamycin treatment. This gene is YBR238C.

Uncharacterized genes mapping unravels the role of YBR238C in cellular aging

Not much is known about YBR238C besides its effect on lifespan (to increase CLS and RLS), its mitochondrial localization 39, its transcriptional up-regulation by TORC1 and the existence of the paralogue RMD9. The encoded protein has 731 amino acid residues. Sequence architecture analysis with the ANNOTATOR 40 reveals an intrinsically unstructured region over the first ca. 130 residues (first, a long polar but uncharged run followed by a histidine/asparagine-rich region beginning with position 83) and a pentatricopetide repeat region (residues 130-675, for example, due to a HHpred 41 hit to structure 7A9X chain A with E-value 2.e-56). Given the sequence homology, we hypothesize that the protein encoded by YBR238C is involved in RNA binding as its paralogue RMD9 with a similar globular segment42.

To note, YBR238C carries the conflicting annotations in SGD for increased and decreased RLS upon deletion. Unfortunately, there is not a single direct report about YBR238C listed in the scientific literature at the time of writing. There are a few genome-wide deletion strain studies that identified YBR238C as one of the gene that increases CLS 43 and RLS 8,44–46. However, the SGD database wrongly documented one of the latter studies as evidence for a decreased RLS phenotype of YBR238C 46. Thus, this examination allows us to claim YBR238C as the only uncharacterized rapamycin downregulated gene causing CLS and RLS increase upon deletion.

First, we confirmed that YBR238C is indeed a rapamycin response gene by qRT-PCR expression analysis in both yeast backgrounds BY4743 and CEN.PK (Figures 1A and 1B). Further, we tested the role of YBR238C in cellular aging by measuring the CLS in yeast. CLS was analysed in BY4743 and CEN.PK strains using three different outgrowth survival methods (see detail in methods section). Cell survival of wild type and ybr238cΔ strains was analysed at various age time points. We found higher CLS of ybr238cΔ cells compared to wild type cells (Figures 1C-1F). Together, these results confirmed that rapamycin inhibits the expression of YBR238C, and deletion of this gene indeed increases the cellular lifespan.

YBR238C deletion increases the cellular lifespan

(A and B) Expression analysis of YBR238C gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (A) and CEN.PK113-7D (B). Data are represented as means ± SD (n=4). ****P < 0.0001 based on two-sided Student’s t-test.

(C and D) Chronological lifespan (CLS) of the wild type and ybr238cΔ mutant was assessed in the SD medium supplemented with auxotrophic amino acids for BY4743 (C) and only the SD medium for CEN.PK113-7D (D) strains in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=3) (C) and (n=4) (D). **P < 0.01, ***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Šídák’s multiple comparisons test.

(E and F) CLS of the wild type and ybr238cΔ mutant was performed in the flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (E) and YPD medium in the 96-well plate (F). The serial outgrowth of aged cells on agar medium was imaged (E) and quantified the survival relative to outgrowth of day 2 (F). A representative of two experiments for (E) and (F) is shown.

Transcriptomics analysis reveals the longevity gene expression signatures of ybr238cΔ mutants

The transcriptome of the long-lived ybr238cΔ mutant was compared with that of the wild type (Figure S3A). Applying standard significance criteria, we found 322 genes up– and 56 genes down-regulated in the ybr238cΔ mutant compared to wild type (Figure S3B; Additional File 3). Thus, the transcriptome of the ybr238cΔ mutant is very distinct from that of the wild type.

Notably, we see several major metabolic changes. For the ybr238cΔ mutant, genes in mitochondrial metabolic processes such as oxidative phosphorylation and aerobic respiration show predominant enrichment (Figures 2A and 2B; Additional File 3). Since YBR238C expression is regulated by TORC1 (Figures 1A and 1B), it is not surprising that we observe similarities in the profiles of upregulated DEGs for the ybr238cΔ mutant and for the case of treating the wild type with rapamycin (Figures S3C and S3D).

Longevity signatures of ybr238cΔ mutant

(A and B) Functional enrichment analysis of upregulated genes in yeast Saccharomyces cerevisiae ybr238cΔ mutant compared to wild type (BY4743). Bar plots showing the enriched biological process (A) and identified MCODE complexes based on the top-three ontology enriched terms (B). See also Additional File 3.

(C and D) ATP analysis of wild type and ybr238cΔ mutant for BY4743 (C) and CEN.PK113– 7D (D). Data are represented as means ± SD (n=6). **P < 0.01, and ****P < 0.0001 based on two-sided Student’s t-test.

(E and F) Expression analysis of MSN4 gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (E) and CEN.PK113-7D (F). Data are represented as means ± SD (n=2). **P < 0.01 based on two-sided Student’s t-test.

(G and H) ROS analysis of wild type and ybr238cΔ mutant for BY4743 (G) and CEN.PK113– 7D (H). Data are represented as means ± SD (n=4). ***P < 0.001, and ****P < 0.0001 based on two-sided Student’s t-test.

Next, we experimentally tested whether the transcriptome longevity signatures is associated with enhanced mitochondrial metabolism, whether the cellular energy level has gone up and cellular stress responses are induced with a switch to oxidative metabolism 47,48. Indeed, our metabolic analysis revealed an increased ATP level in ybr238cΔ mutants compared to wild type cells (Figures 2C and 2D). Hence, the ybr238cΔ mutation rewires the cellular metabolism to promote resource-saving ways of energy production as the up-regulated expression of OXPHOS machinery subunits truly boosts ATP synthesis in ybr238cΔ mutant cells.

The enhanced mitochondrial function in ybr238cΔ mutants does also improve the protection against reactive oxygen species (ROS). Among the 150 TFs that control the upregulated DEG of the ybr238cΔ mutant (Additional File 3), 13 TFs are significantly overrepresented within the upregulated DEGs (Additional File 3). Importantly, we found that the stress response controlling transcription factor MSN4 is up-regulated in ybr238cΔ mutants (Figures 2E and 2F; S4A-S4B) 16,49. Concomitantly, we found less ROS level in ybr238cΔ mutant compared to wild type cells (Figures 2G and 2H).

Notably, the transcription factor regulated activation of stress response pathways (thioredoxins, molecular chaperones, etc.) 50,51 and the switch from fermentation to respiration are associated with delayed cellular aging 47,48,52,53. Our results show that the ybr238cΔ mutation triggers oxidative metabolism and stress protective machineries activation and, as a result increases CLS. Collectively, our findings reveal that YBR238C is a TORC1 regulated gene involved in mitochondrial function coupled with cellular aging. Therefore, we refer to YBR238C as AAG1: Aging Associated Gene 1.

YBR238C negatively regulates mitochondrial function via HAP4-dependent and –independent mechanisms

HAP4 is a transcription factor that controls the expression of mitochondrial electron transport chain components including OXPHOS genes 49. HAP4 activity has been shown to increase lifespan by enhancing the mitochondrial respiration in cells 49. Intriguingly, HAP4 is among the 13 TFs overrepresented among the upregulated DEGs in ybr238cΔ mutant (Figures 3A-3C; Additional File 3). Consistent with these findings, transcription profile of mitochondrial genes in the ybr238cΔ mutant is opposite to that of the hap4Δ mutant (Figure 3D). For example, ETC complexes I – V genes’ expression is increased in ybr238cΔ, however, it is decreased by HAP4 deletion (Figure 3D).

YBR238C affects cellular aging via HAP4-dependent and independent mechanisms

(A) The HAP4 regulon within the upregulated DEGs of ybr238cΔ mutant. See also Additional File 3.

(B and C) Expression analysis of HAP4 gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (B) and CEN.PK113-7D (C). Data are represented as means ± SD (n=2). **P < 0.01 based on two-sided Student’s t-test.

(D) Heatmap of mitochondrial ETC genes expression analysis by qRT-PCR in CEN.PK113– 7D strains. Data are represented as means ± SD (n=2).

(E, F and H) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=6). *P < 0.05, ***P < 0.001, ****P < 0.0001 and ns: non-significant. Ordinary one–way ANOVA followed by Tukey’s multiple comparisons test (E and F). Two-way ANOVA followed by Šídák’s multiple comparisons test (H).

(G) Expression analysis of HAP4 and ETC genes by qRT-PCR in YBR238C deletion (ybr238cΔ) and overexpression (YBR238C-OE) strains of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D (C). Data are represented as means ± SD (n=2). **P < 0.01, ***P < 0.001 and ****P < 0.000 based on Two-way ANOVA followed by Šídák’s multiple comparisons test.

(I) Model representing regulation of cellular aging by YBR238C via HAP4–dependent and independent mechanisms.

We found that deletion of HAP4 moderately decreases CLS at the background of the ybr238cΔ mutant (Figure 3E). To confirm that the decrease in lifespan is through the HAP4 pathway, we examined the expression of mitochondrial respiratory genes. We found that HAP4 deletion significantly decrease the ETC complex I – V genes’ expression in ybr238cΔ mutant (Figure 3D). To note, HAP4 deletion at the wild-type background decreases the cellular lifespan even more (Figure 3E), which is consistent with more dramatic reduction of ETC complexes I – V genes’ expression (Figure 3D). Taken together these data suggests that YBR238C negatively regulates the HAP4 activity. HAP4-upregulated, increased mitochondrial function contributes to the prolonged lifespan of ybr238cΔ mutants.

YBR238C gene deletion rescues some loss of lifespan of hap4Δ mutants (Figure 3F). The effect is especially pronounced until day 8 when wild-type cells are essentially 100% surviving. Yet, complete epistasis of phenotypes is not achieved. This observation parallels the above findings that HAP4 deletion in ybr238cΔ mutant does not fully recover the mitochondrial ETC complexes I – V genes’ expression and CLS at day 9 compared to the wild-type background (Figures 3D and 3E). Together, these results indicate that YBR238C affects cellular lifespan via HAP4-dependent and –independent mechanisms.

To confirm the existence of HAP4-independent mechanisms, we examined the lifespan and transcriptome under conditions of YBR238C overexpression (YBR238C-OE). As expected, YBR238C-OE decreases the expression of HAP4, of mitochondrial ETC complexes I – V genes (Figure 3G) as well as the CLS (Figure 3H). Strikingly, the lifespan of YBR238C-OE cells was shorter than for hap4Δ mutants (Figure 3H). Thus, HAP4-independent mechanism does exist through which YBR238C also affects cellular aging (Figure 3I).

The YBR238C paralogue RMD9 deletion decreases the lifespan of cells

In yeast S. cerevisiae, YBR238C has a paralogue RMD9 that shares ∼45% amino acid identity54. Given that YBR238C activity is tightly coupled with the lifespan, we investigated the role of its paralogue RMD9 in cellular aging. Surprisingly, we found that RMD9 deletion has an effect opposite to YBR238C deletion and it shortens CLS (Figures 4A-4C; S5A-S5C).

RMD9 deletion decreases the cellular lifespan

(A and E) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=12) (A) and (n=6) (B). ****P < 0.0001 and ns: non-significant. Ordinary one-way ANOVA followed by Tukey’s multiple comparisons test (A). Two-way ANOVA followed by Dunnett’s multiple comparisons test (E).

(B, C, F and G) CLS of yeast strains was performed in the flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (B and F) and YPD medium in the 96-well plate (C and G). The serial outgrowth of aged cells on agar medium was imaged (B and F) and quantified the survival relative to outgrowth of day 1 (C and G). A representative of two experiments for B, C, F and G is shown.

(D) Serial dilution growth assays of the yeast strains on fermentative glucose medium (YPD) and respiratory glycerol medium (YPG).

(H) ATP level in yeast strains. Data are represented as means ± SD (n=3). *P < 0.05, and ***P < 0.001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test. See also Figure S5.

RMD9 is known to control the mitochondrial metabolism gene expression by stabilizing and processing of the mitochondrial mRNAs 42,54. We asked whether RMD9 deletion induces mitochondrial dysfunction that, therefore, causes accelerated cellular aging. We first tested the mitochondrial activity by allowing mutant cells to grow under respiratory conditions 55–57. We found that rmd9Δ mutants could perform on glucose media but they failed to grow on glycerol as a carbon source (known to require functional mitochondria; Figures 4D; S5D). Our results are in line with previously observed dysfunctional mitochondria in rmd9Δ mutants 42,54.

In control experiments, we found a similar growth defect phenotype on glycerol medium for cells deficient in PET100 and COX6 mitochondrial genes (Figures 4D; S5D) 57. CLS of pet100Δ and cox6Δ mutants were reduced compared to wild type (Figures 4E-4G; S5E-S5G). We also found lowered ATP– and higher ROS-levels in rmd9Δ, pet100Δ, cox6Δ mutants compared to wild type cells (Figures 4H and S5H).

In contrast, long lived ybr238cΔ mutants efficiently grow on respiratory medium with high ATP– and low ROS-levels (Figures 4D and 4H; S5D and S5H). So far, the results indicate that deletion of YBR238C potentiates the mitochondrial function that, in turn, leads to CLS increase. Thus, YBR238C and its paralogue RMD9 antagonistically affect mitochondrial function, CLS and cellular aging.

YBR238C affects the cellular lifespan via an RMD9-dependent mechanism

Previously, YBR238C deletion was shown to increase the CLS via HAP4-dependent and –independent mechanisms (Figure 3) and to rescue some loss of CLS of the hap4Δ mutant by a HAP4-independent mechanism (Figures 3E and 3F). Here, we ask whether YBR238C deletion suppresses the shortened lifespan of rmd9Δ mutants. We examined the lifespan of double deletion rmd9Δ ybr238cΔ mutant and compared it with rmd9Δ. We found that the deletion of YBR238C largely failed to recover the lifespan of rmd9Δ mutant to wild type cells; however, it partially prevented their early cell death (Figures 5A-5C). These results show that YBR238C deletion increases the cellular lifespan via RMD9-dependent mechanisms. Also, we found that the YBR238C deletion results in increased ATP– and reduced ROS-levels in the rmd9Δ mutant (Figures 5D and 5E).

YBR238C affects cellular aging via RMD9– dependent mechanism

(A, B, G and H) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in the SD medium in the flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (A and G) and YPD medium in the 96-well plate (B and H). The serial outgrowth of aged cells on agar medium was imaged (A and G) and quantified the survival relative to outgrowth of day 2 (B and H). A representative of two experiments for A, B, G and H is shown.

(C and I) CLS of yeast strains was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=6) (C) and (n=12) (I). ****P < 0.0001 based on two-way ANOVA followed by Tukey’s multiple comparisons test (C) and Dunnett’s multiple comparisons test (I). ns: non-significant.

(D and J) ATP level in yeast strains. Data are represented as means ± SD (n=4). *P < 0.05, ***P < 0.001, and ****P < 0.0001 based on on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

(E and K) ROS level in yeast strains. Data are represented as means ± SD (n=4). **P < 0.01, ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test. ns: non-significant.

(F) Expression analysis of RMD9 gene by qRT-PCR in yeast strains. Data are represented as means ± SD (n=4). ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

(L) Effect of antimycin A (AMA) treatment on CLS of yeast strain was assessed in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. A representative heatmap data of three experiments is shown.

(M) Model representing regulation of cellular aging by YBR238C via HAP4 and RMD9 dependent mechanisms.

Intriguingly, we find RMD9 expression up-regulated in ybr238cΔ and down-regulated in YBR238C-OE cells, respectively (Figure 5F). So, we asked whether RMD9 expression change is transcriptionally coupled in cellular lifespan phenotypes. Remarkably, RMD9 overexpression increases the lifespan of cells (Figures 5G-5I) as we would have predicted from the observed changes with the YBR238C deletion phenotype.

To know whether this CLS increase is contributed by enhanced mitochondrial function, we quantified the ATP level in wild type, YBR238C-OE, and RMD9-OE cells. ATP level is higher in RMD9-OE cells than wild-type cells, a result in line with RMD9 positively regulating mitochondrial activity (Figure 5J). Consistent with the above findings, YBR238C overexpression decreases the ATP level (Figure 5J). Next, we assessed the oxidative stress and found that the ROS level in RMD9-OE cells was comparable to wild type. Yet, YBR238C overexpression increases the ROS level (Figure 5K).

It can be shown that the effects of YBR238C and RMD9 are at least partially realized via the mitochondrial ETC pathway. Antimycin A (AMA) is an inhibitor of the ETC complex III, decreasing ATP synthesis (Figure S6A) 58. Also, AMA treatment reduces the lifespan of cells (Figure S6B), confirming that mitochondrial energy supply is critical to delay cellular aging. Since YBR238C and RMD9 affect the ATP level, we tested the AMA effect on their deletion and overexpression strains. We found that ybr238cΔ and RMD9-OE cells with their enhanced mitochondrial function phenotype were largely resistant to AMA treatment (Figure 5L). AMA aggravates the cellular aging of mitochondrial defective YBR238C-OE cells (Figure 5L). Notably, mitochondrial defective rmd9Δ cells were not further affected by AMA treatment (Figure 5L).

Altogether, our findings reveal that YBR238C affects CLS and cellular aging via modulating mitochondrial function by mechanisms including HAP4– and RMD9-dependent pathways (Figure 5M).

YBR238C connects TORC1 signaling with modulating mitochondrial function and cellular aging

So far, we learned that YBR238C is regulated by TORC1 (Figures 1A and 1B) and it affects cellular lifespan by modulating mitochondrial function. Here, we wish establish the direct connection between the TORC1 signaling and mitochondrial activity. We treated cells with the TORC1 inhibitor rapamycin and found that, subsequently, the ATP level increases in the cells (Figures 6A; S6C). To test whether TORC1 affects mitochondrial function via YBR238C, we analyse the expression profile of mitochondrial ETC genes. As expected, rapamycin supplementation decreased the expression of YBR238C (Figures S6D and S6E) but induced the expression of ETC genes (Figure 6B). Remarkably, rapamycin-induced changes in the expression of ETC genes were largely unaffected in ybr238cΔ cells and reduced in YBR238C– OE cells (Figure 6B). Our results are consistent with the hypothesis that TORC1 regulates the mitochondrial ETC genes via YBR238C.

TORC1-MItochondria-TORC1 (TOMITO) signaling axis regulate cellular aging

(A) ATP analysis of logarithmic phase wild type CEN.PK113-7D yeast cells treated with rapamycin for 1 hour. Data are represented as means ± SD (n=3). ****P < 0.0001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test. See also Figure S5C.

(B) Expression analysis of mitochondrial ETC genes by qRT-PCR. Data are represented as means ± SD (n=2).

(C-G) Chronological lifespan (CLS) of yeast strains with indicated concentrations of rapamycin was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=3). ***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant.

(H) CLS of wild type yeast strain with indicated concentrations of rapamycin and antimycin A (10 µM) was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. Data are represented as means ± SD (n=3). ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non– significant. ns: non-significant.

(I) ATP analysis of stationary phase wild type CEN.PK113-7D yeast cells were incubated with rapamycin (10 nM) and antimycin A (10 µM). Data are represented as means ± SD (n=6). ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test. See also Figure S6H.

(J) Determination of human HEK293 cells survival incubated with indicated concentrations of antimycin A for 6 days. Untreated control is considered 100% cell survial. Data are represented as means ± SD (n=6).

(K) Heatmap for cell survival analysis of human HEK293 cells incubated with indicated concentrations of antimycin A and rapamycin for 6 days. Cell viability at antimycin A and rapamycin combination concentrations was measured relative to their individual treated concentration. A representative data of two experiments is shown.

The strains ybr238cΔ and YBR238C-OE are associated with increased and decreased CLS, respectively. So, we ask whether TORC1 influences cellular aging via YBR238C by modulating mitochondrial function. We examined the effect of rapamycin supplementation on the lifespan of ybr238cΔ and YBR238C-OE cells. Strikingly, addition of rapamycin does not further increase CLS of ybr238cΔ cells. Additionally, anti-aging effect of rapamycin is significantly reduced in YBR238C-OE cells (Figure 6C). These findings are consistent with the rapamycin effect on transcriptomics profiles of ETC genes in ybr238cΔ and YBR238C-OE cells (Figure 6B). Taken together, these results reveal that YBR238C is a downstream effector of TORC1 signaling, connecting mitochondrial function for the regulation of cellular aging.

Mitochondrial dysfunction induces the TORC1 activity that causes accelerated cellular aging

We showed that YBR238C-OE cells are associated with mitochondrial dysfunction and shortened lifespan (Figure 5). Apparently, the accelerated cellular aging phenotype is primarily due to compromised cellular energy level that affects the lifespan. Genetic (ybr238cΔ) and chemical (rapamycin) mediated enhancement of ATP level increases the lifespan of wild type cells validate this conclusion. Notably, the rapamycin supplementation significantly rescued the shortened lifespan of mitochondrial defective YBR238C-OE cells (Figures 6D; and S6F). This observation suggests that, as a reaction on sensing mitochondrial dysfunction, TORC1 is activated which, in turn, leads to accelerated cellular aging. Indeed, we found that cellular growth (a proxy for TORC1 activity) of mitochondria dysfunctional YBR238C-OE cells was resistant to rapamycin inhibition at concentrations that were effective for wild type cells (Figure S6G).

Further, we asked what the effect of TORC1 activity under enhanced mitochondrial function conditions would be. To test this, we assessed the growth of ybr238cΔ cells with rapamycin. We found that the cellular growth of ybr238cΔ cells was reduced by rapamycin compared to wild type cells (Figure S6G). Apparently, TORC1 activity is not signalled downstream under these conditions and rapamycin further aggravates this by TORC1 inhibition. This result suggests that YBR238C affects TORC1 activity via modulating mitochondrial function.

We found a similar pattern of rapamycin effect on growth of mutant cells with enhanced or damaged mitochondrial function. Growth of RMD9-OE cells having enhanced mitochondrial function was sensitive to rapamycin (Figure S6G). Likewise, growth of defective mitochondrial rmd9Δ, pet100Δ and cox6Δ mutants is resistant to rapamycin compared to wild type (Figure S6G). The YBR238C deletion reduced the rapamycin growth resistance phenotype of mitochondrial dysfunctional rmd9Δ (Figure S6G) and hap4Δ cells (Figure S6H), apparently by enabling enhanced mitochondrial function.

Taken together, we see that YBR238C plays a junction role in integrating mitochondrial function and TORC1 signaling.

TORC1 inhibition prevents accelerated cellular aging caused by mitochondrial dysfunction in yeast and human cells

Mitochondrial dysfunction increases the TORC1 activity and, subsequently, causes accelerating cellular aging. We asked whether inhibition of TORC1 activity could prevent accelerating cellular aging even under mitochondrial dysfunction conditions. Previously, we have already shown that rapamycin-mediated TORC1 inhibition partly rescued the shortened lifespan of mitochondrial defective YBR238C-OE cells (Figures 6D; and S6F). We examined other mitochondrial dysfunctional conditions to confirm that suppressive effect of rapamycin is not only specific to YBR238C-OE. We tested defective mitochondrial mutants rmd9Δ, pet100Δ and cox6Δ. In all cases, rapamycin supplementation prevents the accelerated cellular aging (Figures 6E-6G).

Rapamycin supplementation also rescued the AMA-mediated accelerated cellular aging (Figure 6H). We also quantified the ATP level to confirm that the suppressive effect of rapamycin is specific to mitochondrial function. Rapamycin supplementation increases the ATP level of AMA-treated cells (Figures 6I; and S6I), demonstrating that TORC1 inhibition improves mitochondrial function and thus suppresses accelerated cellular aging.

We further verified the connection of TORC1 and mitochondrial function in human cells. First, we examined the effect of AMA on the survival of HEK293 cells. AMA treatment decreases the viability of HEK293 cells (Figure 6J). Cell survival reduction is due to the mitochondrial dysfunction as we see lower ATP levels in AMA treated cells compared to untreated control (Figure S6J). Subsequently, we examined the survival of AMA-treated HEK293 cells with or without rapamycin supplementation. Rapamycin supplementation suppresses AMA-mediated cellular death (Figure 6K). Further rapamycin supplementation increases the ATP level treated cells (Figure S6J).

Our findings convincingly support that TORC1 inhibition suppresses cellular aging associated with mitochondrial dysfunction across species.

Discussion

Identifying lifespan regulators, genetic networks, and connecting genetic nodes relevant to cellular aging provides functional insight into the complexity of aging biology 1,7,11 for the subsequent design of anti-aging interventions. Understanding the mechanism of aging will also require to understand the role of many genes of yet unknown function as YBR238C at the beginning of this work. Unfortunately, the group of uncharacterized genes coding proteins of unknown function has received dramatically decreasing attention during the past two decades19–21.

In this study, we first summarized literature and genome database reports about genes affecting aging in the yeast model (mostly observed in gene deletion screens). We found that AAGs (Aging Associated Genes) make up about one-third of the total yeast genome (many of them are functionally uncharacterized) and can be classified into 15 categories for increase/decrease CLS and/or RLS under various conditions.

We compared the set of AAGs with the group of RRGs (Rapamycin Regulated Genes) and found that about one third of the AAGs is TORC1 activity regulated. Among the functionally uncharacterized genes in this overlap set, the gene YBR238C stands out as the only one that (1) is downregulated by rapamycin and (2) its deletion increases both CLS 43 and RLS 8,44–46. Thus, YBR238C is important among identified rapamycin-downregulated uncharacterized genes in AAG categories as it seems to rationally connect the rapamycin-induced TORC1 inhibition with the increase of both CLS and RLS.

We observed that the CLS of ybr238cΔ cells is higher than for wild type, regardless of various aging experimental conditions tested in this study. Transcriptional analysis of ybr238cΔ cells identified a longevity signature including enhanced gene expression related to mitochondrial energy metabolism and stress response genes 9,49,59–62. In contrast, YBR238C-OE cells display mitochondrial dysfunction leading to decreased lifespan. Together, these results reveal that TORC1 regulates the YBR238C expression which is linked to mitochondrial function and cellular lifespan.

The YBR238C paralogue RMD9 has been previously shown to affect mitochondrial function 42,54. Surprisingly, we observed an antagonism regarding the CLS phenotype of deletion/overexpression for YBR238C and RMD9. YBR238C overexpression and RMD9 deletion confer defective mitochondrial function associated with accelerated cellular aging and decrease lifespan of cells. In contrast, YBR238C deletion and RMD9 overexpression enhance the mitochondrial function that increase the cells’ longevity. Consistent with the identified negative regulatory role of YBR238C on mitochondrial function, deletion of this gene partially suppressed the accelerated aging of rmd9Δ cells and AMA-treated cells. We identified that YBR238C influences mitochondrial function and the CLS phenotype with contributions from HAP4– and RMD9-dependent mechanisms. As (i) one of RMD9’s molecular functions has been shown to stabilize certain (mitochondrial) mRNAs and (ii) YBR238C has a has a homologous pentatricopeptide repeat region, the two paralogues might differ in the sets of protected mRNAs with opposite outcome on cellular longevity.

While Kaeberlein et al. identified YBR238C as a top candidate for increasing replicative lifespan (RLS) in yeast through deletion 45, our study builds upon their work by investigating the mechanisms and the connection with TORC1 in chronological lifespan (CLS). Results of genetic interventions (YBR238C deletion and overexpression) and rapamycin treatment show that YBR238C is a downstream effector by which TORC1 modulates mitochondrial activity. Most importantly, we found that TORC1 inhibition increases the mitochondrial function via YBR238C and, thereby, extends cellular lifespan. TORC1 is involved in both CLS and RLS 16. Whether the TORC1 – YBR238C axis has similar or distinct mechanisms for CLS and RLS will be interesting to identify in the future.

TORC1 is the major controller of cellular metabolism that links environmental cues and signals for cellular growth and homeostasis. TORC1 upregulates anabolic processes such as de novo synthesis of proteins, nucleotides, lipids, and downregulates catabolic processes such as inhibition of autophagy 26,28. We find that dysfunctional mitochondria lead to TORC1 activation both in yeast and in human cells with accompanying onset of accelerated cellular aging. Our results are in line with a recent report about anabolic pathways enhancement and suppression of catabolic processes in cells with defective mitochondria 63. On the other hand, TORC1 activity decreases under enhanced mitochondrial function environment. The cause of the mitochondrial dysfunction (whether due to deletion of a critical gene or pharmacological intervention) is irrelevant in this context. Nevertheless, despite the mitochondrial insufficiency background, inhibition of TORC1 (e.g., with rapamycin) can partially rescue the shortened cellular lifespan (Figures 6D-6H; and S6F).

Our work sheds light onto the questions (i) how mitochondrial dysfunction is linked to accelerated cellular aging and (ii) how TORC1 inhibition suppresses the mitochondrial dysfunction and prevents shortening lifespan. Apparently, mitochondrial dysfunction aberrantly signals to increase the TORC1 activity that leads to accelerated aging in cells. Remarkably, TORC1 inhibition can often suppress the accelerated cellular aging associated with impaired mitochondrial function. Yet, we found that growth of mutant strains with dysfunctional mitochondria can be resistant despite TORC1 inhibition by rapamycin with concentrations effective in wild-type cells possibly due to the dramatic increase of TORC1 activity as in YBR238C-OE, rmd9Δ, pet100Δ, cox6Δ and hap4Δ cells (Figures S6G and S6H).

Our interpretation of the experimental results is supported by two recently published studies: (1) TORC1 activation was observed after mitochondrial ETC dysfunction in an induced pluripotent cell model 64. (2) The inhibition of TORC1 delays the progression of brain pathology of mice with ETC complex I NDUFS4-subunit knockout 65.

We think that it will be insightful to explore TORC1 activity under various conditions with compromised mitochondrial function including mitochondrial fission and fusion dynamics which is reported to affect in cell viability 66,67. Altogether, our findings uncover the central role of the feedback loop between mitochondrial function and TORC1 signaling (TORC1– MItochondria-TORC1 (TOMITO) signaling process). Whereas the effector from TORC1 to mitochondria involves YBR238C, the other direction appears executed via metabolite sensing (e.g., α-keto-glutarate, glutamine, etc.) 68.

Methods

Data acquisition

The gene lists that modulate cellular lifespan in aging model organism yeast Saccharomyces cerevisiae were extracted from database SGD 22 and GenAge 23 (as of 8th November 2022). The actual gene lists are available in ‘Additional File 1’.

Yeast strains, growth media and cell culture

The S. cerevisiae auxotrophic BY4743 (Euroscarf) and prototrophic CEN.PK113-7D 69 strains were used in this study. Deletion strains for BY4743 obtained from yeast homozygous diploid collection. Deletion strains in CEN.PK background was generated using standard PCR-based method 70. Yeast strains were revived from frozen glycerol stock on YPD agar (1% Bacto yeast extract, 2% Bacto peptone, 2% glucose and 2.5% Bacto agar) medium for 2-3 days at 30°C.

Human cell lines, growth media and cell culture

Human embryonic kidney cell line (HEK293, ATCC) was cultured in high-glucose DMEM supplemented with 10% FBS and 1% Penicillin Streptomycin Solution. All cells cultured in a humidified incubator with 5% CO2 at 37°C.

Chemical treatment to cell culture

Stock solution of rapamycin and antimycin A was prepared in dimethyl sulfoxide (DMSO). The final concentration of DMSO did not exceed 1% in yeast and 0.01% in human cell lines experiments.

Yeast aging assay

For the chronological lifespan (CLS) experiments, prototrophic CEN.PK113-7D strains were grown in synthetic defined (SD) medium contain 6.7 g/L yeast nitrogen base with ammonium sulfate without amino acids and 2% glucose. SD medium supplemented with histidine (40 mg/L), leucine (160 mg/L), and uracil (40 mg/L) for auxotrophic BY4743 strains. Chronological aging was assessed by determining the lifespan of yeast as described previously with slight modifications 40. Yeast culture grown in overnight at 30°C with 220 rpm shaking in glass flask was diluted to starting optical density at 600 nm (OD600) ∼ 0.2 in fresh medium to initiate the CLS experiment. CLS was performed by outgrowth method utilizing three different approaches: (i) Cells grown and aged in 96-well plates with a total 200 µL culture in SD medium at 30°C. At various age time points yeast stationary culture (2μL) were transferred to a second 96-well plate containing 200μL YPD medium and incubated for 24 hours at 30°C without shaking. Outgrowth OD600 for each age point was measured by the microplate reader; (ii) Cells grown and aged in flask with a total culture volume more than 5 ml SD medium and incubated for 24 hours at 30°C with 220 rpm shaking. At different age time points yeast stationary culture washed and normalized to OD600 of 1.0 with YPD medium. Further, normalized yeast cells were serial 10-fold diluted with YPD medium in 96-well plates. 3 μL of diluted culture were spotted onto the YPD agar plate and incubated for 48 hours at 30°C. The outgrowth of aged cells on the YPD agar plate was photographed using the GelDoc imaging system; (iii) The above discussed serial 10-fold diluted yeast stationary culture with YPD medium in 96-well plates incubated for 24 hours at 30°C without shaking. Outgrowth OD600 for serial diluted aged cells was measured by the microplate reader.

RNA extraction

Yeast cells were first mechanically lysed using the manufacturer’s disruption protocol. Total RNA from yeast cells was extracted using Qiagen RNeasy mini kit. ND-1000 UV-visible light spectrophotometer (Nanodrop Technologies) and Bioanalyzer 2100 with the RNA 6000 Nano Lab Chip kit (Agilent) was used to assess the concentration and integrity of RNA.

RNA sequencing and bioinformatics analysis

RNA sequencing (RNA-Seq) was conducted using NovaSeq PE150. Raw Fastq files were then passed into Fastp v0.23.2 for adapter trimming and low-quality reads removal. Both single end and paired end reads were processed with default parameters with ––detect adapter for pe. Raw reads that passed the quality check were then aligned using HiSat 2 v2.2.1 with the index built from Saccharomyces cerevisiae R-64-1-1 top-level DNA fasta file obtained from Ensembl for sequencing data with S288C strain background. Library information for all experiments were first checked with RSeQC infer_experiment.py v4.0.0 before the addition of the respective parameters for alignment and counts. The resulting SAM files were then converted and sorted to BAM files using Samtools v1.13. The BAM files were then used to generate feature counts using HTSeq v1.99.2. HTSeq counts from each experiment were then used for downstream Differentially Expressed Genes (DEG) analysis. Counts generated from HTSeq were then used for differential gene expression analysis using EdgeR v3.34.1 quasi likelihood F-test and DESeq2 v1.36.0. Principal Component Analysis (PCA) was conducted via the use of Transcript Per Million normalized counts. Samples that are separated by batches in the PCA were corrected using ComBat-Seq v3.44.0 before PCA was conducted after normalization of the corrected counts. Functional enrichment analysis was performed by metascape tool 28.

qRT-PCR analysis

Real time qRT-PCR experiments were performed as described previously using QuantiTect Reverse Transcription Kit (Qiagen) and SYBR Fast Universal qPCR Kit (Kapa Biosystems) 42. The abundance of each gene was determined relative to the house-keeping transcript ACT1.

ATP analysis

Yeast cells were mixed with the final concentration of 5% trichloroacetic acid (TCA) and then kept on ice for at least 5 min. Cells were washed and resuspended in 10% TCA and lysis was performed with glass beads in a bead beater to extract the ATP. ATP extraction from human HEK293 cells was performed using Triton X-100 lysis buffer. The ATP level was quantified by PhosphoWorks™ Luminometric ATP Assay Kit (AAT Bioquest) and normalized by protein content measured Bio-Rad protein assay kit.

ROS measurement

Cells were washed and resuspended in 1× phosphate buffer saline (PBS, pH 7.4). After that cells were incubated with 40 μM H2DCFDA (Molecular probe) for 30 min at 30°C. Cells were then washed with PBS and ROS level was measured by fluorescence reading (excitation at 485nm, emission at 524nm) by the microplate reader. The fluorescence intensity was normalized with OD600.

Transcription factors analysis

Transcription factors (TFs) enrichment analysis was performed using YEASTRACT 71,72. The significant p-value <0.05 considered for regulatory network analysis based on DNA-binding plus expression evidence. The TFs regulatory networks were visualized with a force-directed layout.

Fermentative and respiratory growth assay

Yeast cells grown in YPD medium were washed and normalized to OD600 of 1.0 with water. Further, normalized yeast cells were serial 10-fold diluted with water. 3 μL of diluted culture were spotted onto the agar medium YPD (2% glucose) and YPG (3% glycerol) and incubated for 48 hours at 30°C. The cell growth on the agar plate was photographed using the GelDoc imaging system.

Growth sensitivity assay

Effect of chemical compounds on cell growth was carried out in 96-well plates. At an OD600 of ∼ 0.2 in SD medium 200 µL yeast cells was transferred into the 96-well plate containing serially double-diluted concentrations of compounds. Cells were incubated at 30°C and the growth was measured at OD600 by the microplate reader.

Quantification and statistical analysis

Data analysis of all the experimental results such as mean value, standard deviations, significance, and graphing were performed using GraphPad Prism v.9.3.1 software. The comparison of obtained results were statistically performed using the Student’s t-tests, Ordinary One-way ANOVA and Two-way ANOVA followed by multiples comparison tests. In all the graph plots, P values are shown as *P < 0.05, **P < 0.01, ***P < 0.001, and ****P < 0.0001 were considered significant. ns: non-significant.

Data availability

Additional File 1, Additional File 2, and Additional File 3 data are deposited.

Lead contact and materials availability

Further information and requests for resources and reagents should be directed to and will be fulfilled by the Lead Contact, Dr. Mohammad Alfatah (alfatahm@bii.a-star.edu.sg).

Supplementary figure legends

Transcriptomics analysis of rapamycin to identify the TORC1 regulated genes

(A and B) Chronological lifespan (CLS) of the wild type BY4743 (A) and CEN.PK113-7D (B) yeast strains was assessed in the SD medium supplemented with auxotrophic amino acids (A) and only the SD medium (B) in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=4). *P < 0.05, and ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non– significant.

(C) Principal component analysis (PCA) of the RNA-seq data from replicates of DMSO and rapamycin (50 nM) treated yeast cells (BY4743). PC1 represents 95.98% of the total variance in the experimental data, and PC2 represents 2.59% of the total variance.

(D and E) Volcano plots produced by EdgeR (D) and DESeq2 (E) demonstrate the gene expression pattern and the differentially expressed genes (DEGs) with p-value <0.05, with a fold change of log2(fold change) >1 for upregulated genes and log2(fold change) <−1 for downregulated genes.

(F) Overlap of identified DEGs using EdgeR and DESeq2 with the criteria for all two tools that p-value <0.05, with a fold change of log2(fold change) >1 for upregulated genes and log2(fold change) <−1 for downregulated genes. Upregulated and downregulated DEGs overlaps for both EdgeR and DESeq2 were further considered for functional GO enrichment analysis.

See also Additional File 2.

RNA sequencing data validation by qRT-PCR

(A and B) Expression analysis of selected genes of RNA sequencing (RNA-seq) data by qRT–PCR in in Saccharomyces cerevisiae genetic backgrounds BY4743 (A) and CEN.PK113-7D (B) wild type cells treated with DMSO and rapamycin (50 nM). Gene expression of rapamycin treated samples were compared with DMSO control. Data are represented as means ± SD (n=4). *P < 0.05, **P < 0.01, ***P < 0.001, and ****P < 0.0001 based on two-sided Student’s t-tests.

Identification of the role of uncharacterized YBR238C gene in cellular aging, Related to Figure 2

(A) Principal component analysis (PCA) of the RNA-seq data from replicates of wild type BY4743 and ybr238cΔ yeast strains. PC1 represents 71.75% of the total variance in the experimental data, and PC2 represents 18.15% of the total variance.

(B) Volcano plots produced by DESeq2 demonstrate the gene expression pattern and the differentially expressed genes (DEGs) with p-value <0.05, with a fold change of log2(fold change) >0.5 for upregulated genes and log2(fold change) <−0.5 for downregulated genes. GO enrichment analysis was performed for upregulated DEGs. See also Additional File 3.

(C and D) Comparison of functional enrichment analysis for upregulated DEGs of ybr238cΔ mutant and rapamycin (50 nM) treated wild type BY4743 cells. Bar plots showing the common enriched biological process (C) and MCODE complexes based on the top-three ontology enriched terms (D).

Transcription factors analysis for upregulated DEGs of ybr238cΔ mutant, Related to Figure 2

(A) The whole regulatory network of overrepresented top eight enriched transcription factors (blue) within the upregulated DEGs of ybr238cΔ mutant. Msn4 and Hap4 TFs are appearing as responsible for the highest number of regulated genes in ybr238cΔ mutant.

(B) The Msn4 regulon within the upregulated DEGs of ybr238cΔ mutant.

See also Additional File 3.

YBR238C homolog RMD9 deletion leads to mitochondrial dysfunction associated with accelerated cellular aging, Related to Figure 4

(A and E) Chronological lifespan (CLS) of yeast wild type BY4743 and deletion strains was performed in the SD medium supplemented with auxotrophic amino acids in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. (A) Data are represented as means ± SD (n=12). ****P < 0.0001 based on two-way ANOVA followed by Tukey’s multiple comparisons test. (E) Data are represented as means ± SD (n=4). ****P < 0.0001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant.

(B, C, F and G) CLS of yeast wild type BY4743 and deletion strains was performed in the flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (B and F) and YPD medium in the 96-well plate (C and G). The serial outgrowth of aged cells on agar medium was imaged (B and F) and quantified the survival relative to outgrowth of day 1 (C and G). A representative of two experiments for B, C, F and G is shown.

(D) Serial dilution growth assays of the yeast strains on fermentative glucose medium (YPD) and respiratory glycerol medium (YPG).

(H) ROS level of yeast wild type BY4743 and deletion strains. Data are represented as means ± SD (n=3). **P < 0.01, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test.

TORC1-Mitochondrial function signaling pathways control cellular aging

(A and C) ATP analysis of wild type CEN.PK113-7D yeast cells treated with (A) antimycin A (stationary phase culture) and (C) rapamycin (logarithmic phase culture). Data are represented as means ± SD (n=3). ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test.

(B) Chronological lifespan (CLS) of wild type yeast strain with indicated concentrations of antimycin A was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. Data are represented as means ± SD (n=8). ***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant. ns: non-significant.

(D and E) Expression analysis by qRT-PCR for YBR238C in logarithmic phase yeast strains treated with rapamycin (200 nM) for 1 hour. Data are represented as means ± SD (n=2). **P < 0.01, and ****P < 0.0001 based on two-sided Student’s t-test.

(F) CLS of yeast strains with indicated concentrations of rapamycin was performed in the SD medium in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=3). ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant. See also Figure 6D.

(G) Growth assay of yeast strains with rapamycin (40 nM) for 24 and 48h. Growth was normalized with untreated control.

(H) Growth assay of yeast strains with indicated concentrations of rapamycin for 48, 72 and 96 h. Growth was normalized with untreated control.

(I) ATP analysis of stationary phase wild type CEN.PK113-7D yeast cells were incubated with indicated concentrations of rapamycin. Data are represented as means ± SD (n=6). ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

(J) ATP analysis of human HEK293 cells treated with antimycin A (10 µM) and rapamycin (100 nM) for 48 hours. Data are represented as means ± SD (n=3). ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

Funding

This work was supported by Bioinformatics Institute (BII), A*STAR Career Development Fund (C210112008) and the Global Healthy Longevity Catalyst Awards grant (MOH-000758– 00).

Author contributions statement

Mohammad Alfatah: Conceiving of the project, Writing, Editing and Funding Acquisition Jolyn Lim Jia Jia: Methodology, Investigation, Bioinformatics, Formal analysis, and Reviewing

Yizhong Zhnag: Methodology, Investigation, Formal analysis, and Reviewing

Arshia Naaz: Methodology, Formal analysis, Investigation and Reviewing

Trishia Cheng Yi Ning: Methodology, Formal analysis, Investigation and Reviewing

Sonia Yogasundaram: Methodology, Investigation and Reviewing

Nashrul Afiq Faidzinn: Methodology, Investigation and Reviewing

Jovian Lin Jing: Methodology, Investigation and Reviewing

Birgit Eisenhaber: Methodology, Investigation, Bioinformatics, Formal analysis and Reviewing

Frank Eisenhaber: Investigation, Bioinformatics, Formal analysis, Reviewing, Editing and Funding Acquisition

Supporting information

eLife Assessment

This valuable study identifies an uncharacterized yeast gene regulating chronological lifespan in a mitochondrial-dependent pathway. The approach to identify and characterise this new gene is compelling, but the evidence is incomplete in supporting the major conclusions. With a stronger focus on the relevance of replicative in addition to chronological lifespan, and stronger data linking to mitochondrial function, this paper would be interesting to the yeast biologists working in metabolism and aging.

Reviewer #1 (Public Review):

Summary

This fascinating paper by M. Alfatah et al. describes work to uncover novel genes affecting lifespan in the budding yeast S. cerevisiae, eventually identifying and further characterizing a gene, YBR238C, now named AAG1 by the authors.

The authors began by considering published gene sets pulled from the Saccharomyces genome database that described increases or decreases in either chronological lifespan or replicative lifespan in yeast. They also began with gene sets known to be downregulated upon treatment with the lifespan-extending TOR inhibitor rapamycin.

YBR283C was unique in being largely uncharacterized, downregulated upon rapamycin treatment, and linked to both increased replicative lifespan and increased chronological lifespan upon deletion.

The authors show that YBR283C may act to negatively regulate mitochondrial function, in ways that are both dependent on and independent of the stress-responsive transcription factor Hap4, largely by looking at relative expression levels of relevant mitochondrial genes.

In a hard-to-fully interpret but well-documented series of experiments the authors note that the two paralogues YBR283C and RMD9 (which have ~66% similarity) (a) have opposite effects when acting alone, and (b) appear to interact in that some phenotypes of ybr283c are dependent on RMD9.

A particularly interesting finding in light of the current literature and of the authors' strategy in identifying YBR283C is that changes in electron transport chain genes upon rapamycin treatment appear to be affected via YBR283C.

Based on a series of experiments the authors move to conclude the existence of "a feedback loop between TORC1 and mitochondria (the TORC1-Mitochondria-TORC1 (TOMITO) signaling process) that regulates cellular aging processes."

Strengths

Overall, this study describes a great deal of new data from a large number of experiments, that shed light on the potential specific roles of YBR238C and its paralog RMD9 in aging in yeast, and also underscore the potential of an approach looking for "dark matter" such as uncharacterized genes when seining the increasing deluge of published datasets for new hypotheses to test. This work when revised will become a valuable addition to the field.

Weaknesses

A paralog of YBR283C, RMD9, also exists in the yeast genome. While the authors indicate that part of their interest in YBR283C lies in its uncharacterized nature, its paralogue, RMD9, is not uncharacterized but is named due to its phenotype of Required for Meiotic nuclear Division, which is not mentioned or discussed anywhere in the manuscript currently.

In the context of the current work, in addition to the cited Hillen, H.S et al. and Nouet C. et al, the authors might be very interested in the 2007 Genetics paper "Translation initiation in Saccharomyces cerevisiae mitochondria: functional interactions among mitochondrial ribosomal protein Rsm28p, initiation factor 2, methionyl-tRNA-formyltransferase and novel protein Rmd9p" (PMID: 17194786), which does not appear to be cited or discussed in the current version of the manuscript.

Reviewer #2 (Public Review):

The effectors of cellular aging in yeast have not been fully elucidated. To address this, the authors curated gene expression studies to link genes influenced by rapamycin - a well-known mediator of longevity across model systems - to genes known to affect chronological and replicative lifespan (RLS) in yeast. Through their analyses, they find one gene, ybr238c, whose deletion increases both CLS and RLS upon deletion and that is downregulated by rapamycin. Curiously, despite these selection criteria, the authors only use CLS as a proxy for cellular aging throughout their study and do not explore the effects of ybr238c deletion on RLS. This does not diminish their conclusions, but given the importance of this phenotype in their selection criteria, it is surprising that the authors did not choose to test both types of aging throughout their study.

Nonetheless, the authors demonstrate that deletion of ybr238c increases CLS across multiple yeast strains and through multiple assays. The authors also test the effects of YBR238C overexpression on lifespan and find the opposite effect, with overexpression yeast showing decreased survival relative to wild-type cells, consistent with "accelerated aging" as the authors propose. The authors also note that ybr238c has a paralog, rmd9, whose deletion decreases CLS and seems to be epistatic to ybr238c, as a double ybr238c/rmd9 mutant has decreased CLS relative to a wild-type strain.

Collectively, the data presented by the authors convincingly demonstrate that ybr238c influences lifespan in a manner that is distinct from (and likely opposite to) rmd9. However, the authors then link the increased CLS in Δybr238c yeast to mitochondrial function using only a handful of assays that do not directly test mitochondrial function. These include total cellular ATP levels, levels of reactive oxygen species, and the transcript levels of select nuclear-encoded mitochondrial genes. Yeast is well established to generate ATP through non-mitochondrial pathways such as glycolysis in fermentive conditions. While it is possible that the ATP levels assayed in the manuscript were tested in stationary phase, which would more likely reflect "mitochondrial function," the methods nor the figure legends contain these details, which are critical for the interpretation of these data. Similarly, ROS can be generated through non-mitochondrial pathways, and the transcription of nuclear-encoded mitochondrial genes is an indirect measure of mitochondrial function at best. Thus, the authors' proposed connection of ybr238c to mitochondrial function is correlative and should be substantiated with assays that more closely align with organellar function, such as respirometry or assaying the activity of oxidiative phosphorylation complexes. Finally, the authors attempt to tie the phenotypes of mitochondrial dysfunction caused by the deletion of ybr238c to TORC1 signaling, as the gene is influenced by rapamycin. However, the presentation of the data, such as reporting ATP levels as relative percentages or failing to perform appropriate statistical comparisons between conditions in which the authors derive conclusions, renders the data difficult to interpret. As such, this manuscript establishes that ybr238c is rapamycin responsive and influences CLS, but its influence on mitochondrial activity and ties to TORC1 signaling remain speculative.

Reviewer #3 (Public Review):

Summary:

The study by Alfatah et al. presented a role for YBR238C in mediating lifespan through improved mitochondrial function in a TOR1-dependent metabolic pathway. The authors used a dataset comparison approach to identify genes positively modulating yeast chronological (CLS) and Replicative (RLS) lifespan when deleted, and their expression is reduced under Rapamycin treatment condition. This approach revealed an unknown, mitochondria-localized yeast gene YBR238C, and through mechanistic studies, they identified its paralogous gene RMD9 regulating lifespan in an antagonistic effect.

Strengths:

Findings have valuable implications for understanding the YBR238C-mediated, mitochondrial-dependent yeast lifespan regulation, and the interplay between two paralogous genes in the regulation of mitochondrial function represents an inserting case for gene evolution.

Weaknesses:

Overall, the implication/findings of this study are restricted only to the yeast model since these two genes do not have any homology in higher eukaryotes. The primary methods must be carefully designed by considering two different metabolic states: respiration-associated with CLS and fermentation-associated with RLS in a single comparative approach. Yeast CLS and RLS are two completely different processes. It is already known that most gene-regulating CLS is not associated with RLS or vice versa. The method section is poorly written and missing important information. The experimental approaches are poorly designed, and variability across the datasets (e.g., media condition "YPD," "SC" etc.) and their experimental conditions are not well described/considered; thus, presented data are not conclusive, which decreases the overall rigor of the study.

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary

This fascinating paper by M. Alfatah et al. describes work to uncover novel genes affecting lifespan in the budding yeast S. cerevisiae, eventually identifying and further characterizing a gene, YBR238C, now named AAG1 by the authors. The authors began by considering published gene sets pulled from the Saccharomyces genome database that described increases or decreases in either chronological lifespan or replicative lifespan in yeast. They also began with gene sets known to be downregulated upon treatment with the lifespan-extending TOR inhibitor rapamycin.

YBR283C was unique in being largely uncharacterized, downregulated upon rapamycin treatment, and linked to both increased replicative lifespan and increased chronological lifespan upon deletion.

The authors show that YBR283C may act to negatively regulate mitochondrial function, in ways that are both dependent on and independent of the stressresponsive transcription factor Hap4, largely by looking at relative expression levels of relevant mitochondrial genes.

In a hard-to-fully interpret but well-documented series of experiments the authors note that the two paralogues YBR283C and RMD9 (which have ~66% similarity) (a) have opposite effects when acting alone, and (b) appear to interact in that some phenotypes of ybr283c are dependent on RMD9.

A particularly interesting finding in light of the current literature and of the authors' strategy in identifying YBR283C is that changes in electron transport chain genes upon rapamycin treatment appear to be affected via YBR283C.

Based on a series of experiments the authors move to conclude the existence of "a feedback loop between TORC1 and mitochondria (the TORC1-Mitochondria-TORC1 (TOMITO) signaling process) that regulates cellular aging processes."

Strengths

Overall, this study describes a great deal of new data from a large number of experiments, that shed light on the potential specific roles of YBR238C and its paralog RMD9 in aging in yeast, and also underscore the potential of an approach looking for "dark matter" such as uncharacterized genes when seining the increasing deluge of published datasets for new hypotheses to test. This work when revised will become a valuable addition to the field.

Weaknesses

A paralog of YBR283C, RMD9, also exists in the yeast genome. While the authors indicate that part of their interest in YBR283C lies in its uncharacterized nature, its paralogue, RMD9, is not uncharacterized but is named due to its phenotype of Required for Meiotic nuclear Division, which is not mentioned or discussed anywhere in the manuscript currently.

In the context of the current work, in addition to the cited Hillen, H.S et al. and Nouet C. et al, the authors might be very interested in the 2007 Genetics paper "Translation initiation in Saccharomyces cerevisiae mitochondria: functional interactions among mitochondrial ribosomal protein Rsm28p, initiation factor 2, methionyl-tRNAformyltransferase and novel protein Rmd9p" (PMID: 17194786), which does not appear to be cited or discussed in the current version of the manuscript.

Thank you for your thorough and insightful review of our manuscript. We value your positive feedback and recognition of the strengths in our study. Your constructive comments have been carefully considered, leading to the inclusion of RMD9, identified as 'Required for Meiotic Nuclear Division,' and the addition of the relevant reference (PMID: 12586695) in the revised manuscript. This information has been incorporated into the second paragraph of the "The YBR238C paralogue RMD9 deletion decreases the lifespan of cells" results section.

Furthermore, we appreciate the reviewer's suggestion to include the 2007 Genetics paper on translation initiation in Saccharomyces cerevisiae mitochondria (PMID: 17194786). This citation has been integrated into our revised manuscript.

We believe that these revisions significantly strengthen the manuscript and address the concerns raised by Reviewer #1. We thank the reviewer for their time and valuable input.

Reviewer #2 (Public Review):

The effectors of cellular aging in yeast have not been fully elucidated. To address this, the authors curated gene expression studies to link genes influenced by rapamycin - a well-known mediator of longevity across model systems - to genes known to affect chronological and replicative lifespan (RLS) in yeast. Through their analyses, they find one gene, ybr238c, whose deletion increases both CLS and RLS upon deletion and that is downregulated by rapamycin. Curiously, despite these selection criteria, the authors only use CLS as a proxy for cellular aging throughout their study and do not explore the effects of ybr238c deletion on RLS. This does not diminish their conclusions, but given the importance of this phenotype in their selection criteria, it is surprising that the authors did not choose to test both types of aging throughout their study.

Nonetheless, the authors demonstrate that deletion of ybr238c increases CLS across multiple yeast strains and through multiple assays. The authors also test the effects of YBR238C overexpression on lifespan and find the opposite effect, with overexpression yeast showing decreased survival relative to wild-type cells, consistent with "accelerated aging" as the authors propose. The authors also note that ybr238c has a paralog, rmd9, whose deletion decreases CLS and seems to be epistatic to ybr238c, as a double ybr238c/rmd9 mutant has decreased CLS relative to a wild-type strain.

Collectively, the data presented by the authors convincingly demonstrate that ybr238c influences lifespan in a manner that is distinct from (and likely opposite to) rmd9. However, the authors then link the increased CLS in Δybr238c yeast to mitochondrial function using only a handful of assays that do not directly test mitochondrial function. These include total cellular ATP levels, levels of reactive oxygen species, and the transcript levels of select nuclear-encoded mitochondrial genes. Yeast is well established to generate ATP through non-mitochondrial pathways such as glycolysis in fermentive conditions. While it is possible that the ATP levels assayed in the manuscript were tested in stationary phase, which would more likely reflect "mitochondrial function," the methods nor the figure legends contain these details, which are critical for the interpretation of these data. Similarly, ROS can be generated through non-mitochondrial pathways, and the transcription of nuclear-encoded mitochondrial genes is an indirect measure of mitochondrial function at best. Thus, the authors' proposed connection of ybr238c to mitochondrial function is correlative and should be substantiated with assays that more closely align with organellar function, such as respirometry or assaying the activity of oxidiative phosphorylation complexes. Finally, the authors attempt to tie the phenotypes of mitochondrial dysfunction caused by the deletion of ybr238c to TORC1 signaling, as the gene is influenced by rapamycin. However, the presentation of the data, such as reporting ATP levels as relative percentages or failing to perform appropriate statistical comparisons between conditions in which the authors derive conclusions, renders the data difficult to interpret. As such, this manuscript establishes that ybr238c is rapamycin responsive and influences CLS, but its influence on mitochondrial activity and ties to TORC1 signaling remain speculative.

We would like to express our gratitude to Reviewer #2 for the thoughtful feedback on our manuscript. We have carefully considered your comments and have made comprehensive revisions to address the concerns raised.

We appreciate the suggestion to investigate the role of YBR238C in replicative lifespan (RLS). However, we want to bring to your attention that four previous studies (references 7, 39, 40, and 41) have already identified the involvement of YBR238C in the RLS phenotype. Given the existing body of literature on this aspect, we chose not to duplicate these efforts in our study.

Instead, we focused our efforts on validating the role of YBR238C in chronological lifespan (CLS) phenotype, a finding reported in only one genome-wide study (reference 38). To enhance the comprehensiveness of our study, we performed analyses on different phenotypes, including mitochondria activity and oxidative stress, under both logarithmic-phase (condition for RLS) and stationary phase (condition for CLS). We now clearly indicate the logarithmic-phase/stationary phase conditions in the figure legends of the manuscript, specifying whether the conditions are relevant to RLS or CLS. Additional results of the new experiments have been included in the revised manuscript as supplementary figures (S3E-S3I).

To address concerns about the indirect nature of our mitochondrial function assays, we have performed relative mitochondria content (S3F), quantification of ROS levels from fermentative to stationary phase conditions (S3G), and assessment in respiratory glycerol medium (S3H), which provides a more direct insight into mitochondrial biology. Additionally, we have investigated the resistance of ybr238c∆ cells to H2O2 toxicity and found them to be more resistant compared to wild-type cells.

We believe these revisions strengthen the scientific rigor and clarity of our study. We sincerely appreciate the guidance from Reviewer #2, and we hope these modifications address the concerns raised effectively.

Reviewer #3 (Public Review):

Summary: The study by Alfatah et al. presented a role for YBR238C in mediating lifespan through improved mitochondrial function in a TOR1-dependent metabolic pathway. The authors used a dataset comparison approach to identify genes positively modulating yeast chronological (CLS) and Replicative (RLS) lifespan when deleted, and their expression is reduced under Rapamycin treatment condition. This approach revealed an unknown, mitochondria-localized yeast gene YBR238C, and through mechanistic studies, they identified its paralogous gene RMD9 regulating lifespan in an antagonistic effect.

Strengths:

Findings have valuable implications for understanding the YBR238C-mediated, mitochondrial-dependent yeast lifespan regulation, and the interplay between two paralogous genes in the regulation of mitochondrial function represents an inserting case for gene evolution.

Weaknesses:

Overall, the implication/findings of this study are restricted only to the yeast model since these two genes do not have any homology in higher eukaryotes. The primary methods must be carefully designed by considering two different metabolic states: respiration-associated with CLS and fermentation-associated with RLS in a single comparative approach. Yeast CLS and RLS are two completely different processes. It is already known that most gene-regulating CLS is not associated with RLS or vice versa. The method section is poorly written and missing important information. The experimental approaches are poorly designed, and variability across the datasets (e.g., media condition "YPD," "SC" etc.) and their experimental conditions are not well described/considered; thus, presented data are not conclusive, which decreases the overall rigor of the study.

We sincerely appreciate your thorough review of our manuscript and your insightful comments. We acknowledge the limitation of our study being yeast-specific due to the absence of homologous genes in higher eukaryotes. However, we would like to highlight the significance of our findings in revealing a feedback loop between mitochondrial function and TORC1 signaling (TORC1-Mitochondria-TORC1 or TOMITO signaling process) in cellular lifespan regulation.

Our interpretation of the experimental results is grounded in recent literature. Two studies (references 62 and 63) support our findings by demonstrating TORC1 activation after mitochondrial electron transport chain dysfunction and the delay in brain pathology progression upon TORC1 inhibition, respectively. These studies, discussed in our manuscript, reinforce the relevance of our work in a broader biological context.

We recognize the importance of carefully designing our primary methods to account for the different metabolic states associated with cellular processes, such as respiration in cellular lifespan (CLS) and fermentation in replicative lifespan (RLS). We want to bring to your attention that four previous studies (references 7, 39, 40, and 41) have already identified the involvement of YBR238C in the RLS phenotype. To avoid duplicating these efforts, we have chosen not to reiterate these findings in our study. However, we have clarified the logarithmic-phase/stationary phase conditions in the figure legends, specifying their metabolic states relevance to RLS or CLS. Additionally, we have included new supplementary figures (S3E-S3I) to provide further details on the new experiments conducted.

We appreciate your feedback regarding the clarity and completeness of our method section. In the revised manuscript, we have invested additional effort to enhance the clarity of the method section, providing a more detailed account of the experimental procedures, including the missing information you identified.

We believe these revisions strengthen the scientific rigor and clarity of our study. We sincerely appreciate the guidance from Reviewer #3, and we hope these modifications address the concerns raised effectively.

Reviewer #1 (Recommendations For The Authors):

Thank you for your detailed review and valuable recommendations. We have carefully addressed each of your comments in the revised manuscript. The specific changes made include:

(1) "TORC1 positively regulates aging, and its inhibition increases lifespan in various eukaryotic organisms including yeast and mammalian 13,26,27,29,30." Here I would suggest replacing "mammalian" with "mammals".

We have amended the sentence as recommended.

(2) "Next, we experimentally tested whether the transcriptome longevity signatures are associated with enhanced mitochondrial metabolism, whether the cellular energy level has gone up and cellular stress responses are induced with a switch to oxidative metabolism 47,48." Here I would replace "transcriptome longevity signatures is" with "transcriptome longevity signatures are".

We have amended the sentence as recommended.

(3) "Thus, HAP4-independent mechanism does exist through which YBR238C also affects cellular aging (Figure 3I)." I would replace "Thus, HAP4-independent" with "Thus, a HAP4-independent".

We have amended the sentence as recommended.

(4) "We examined other mitochondrial dysfunctional conditions to confirm that suppressive effect of rapamycin is not only specific to YBR238C-OE." I would change "that suppressive effect" to "that the suppressive effect".

We have amended the sentence as recommended.

(5) "Understanding the mechanism of aging will also require to understand the role of many genes of yet unknown function as YBR238C at the beginning of this work." I would switch "require to understand" to "require understanding".

We have amended the sentence as recommended.

(6) "The gene lists that modulate cellular lifespan in aging model organism yeast Saccharomyces cerevisiae were extracted from database SGD 22 and GenAge 23 (as of 8th November 2022)" "yeast" should not be italicized.

Corrected.

(7) Figure 1, panels C and D, ybr238c should be italicized.

Corrected.

(8) Figure 2B, top left-most (oxidative phosphorylation) network. I might consider repositioning some labels to make them more readable if possible.

Thank you for your feedback. The figure labels in Figure 2B are default from Metascape analysis, so repositioning isn't feasible. However, we have indicated in the figure legends that the full set of genes for functional enrichment analysis and the MCODE complex is available in Additional File 3.

(9) Figure 4E, rmd9, pet100, and cox6 should be italicized.

Corrected.

(10) Figure 5C, rmd9 and rmd9 ybr238c should be italicized. Corrected.

Reviewer #2 (Recommendations For The Authors):

Thank you for your detailed review and valuable recommendations. We have carefully addressed each of your comments in the revised manuscript. The specific changes made include:

(1) The presentation of data as heatmaps (Figures 1F, 3D, 4C, 4G, 5B, 5H, 5L, 6K) obfuscates the quantitative nature of the data. These data would be much stronger if presented as bar graphs with appropriate statistical analysis. If the authors prefer the visual of the heat map, there should be some statistical analysis performed to accompany these figures. This is particularly important for Figure 3D, in which the authors state "We found that HAP4 deletion significantly decrease the ETC complex I-V genes' expression" (bottom of page 8). As no statistical analyses were performed, the authors should refrain from using such language as it is unsupported by the data as analyzed.

Thank you for your insightful comments and suggestions regarding the presentation of our data. We appreciate the attention you have given to Figures 1F, 3D, 4C, 4G, 5B, 5H, 5L, and 6K.

In response to your feedback, we have carefully re-evaluated our approach. Considering the large volume of data associated with our lifespan analysis at different time points, we initially chose to visualize it using heatmaps to comprehensively capture the complexity of the results. However, we have now incorporated quantification information into the heatmaps.

For Figure 3D, which addresses the impact of HAP4 deletion on the expression of ETC complex I-V genes, we have replaced the heatmap with a bar graph. This modification allows for a clearer representation of the quantitative nature of the data. Moreover, we have conducted thorough statistical analyses comparing data between ybr238c∆ and ybr238c∆ hap4∆ to support the statements made in the text. The results of these analyses are now included in the revised figure. Moreover, we also replaced the Figure 6K heatmap with a bar graph.

We believe that these changes enhance the interpretability and robustness of our findings. We are grateful for your guidance, and we are confident that these adjustments will strengthen the overall quality of our manuscript.

(2) The presentation of ATP data, given its importance in supporting the core conclusions of this manuscript, is poor. The conditions under which yeast was collected are not reported, making these data impossible to interpret; total cellular ATP levels would be significantly altered and influenced by separate pathways in fermentive versus stationary phases. Minimally, the authors should describe the conditions of yeast growth (e.g., age, culture media) in which these measurements were made. The presentation of relative ATP percentages is problematic, particularly with measurements that deviate so far from wild-type ATP levels in conditions such as those in Figure 6A, in which the authors report that rapamycin induces a 1200% increase in cellular ATP. Previous papers have established that ATP levels in yeast hover around 4 mM and are stable through the cell cycle and across nutrient conditions (PMID: 30858198, 35438635). Given this, the reported ATP levels would be expected to be near 48 mM, which is strongly outside of the typically accepted values of 1-10 mM for this metabolite. Without understanding the contexts in which these measurements are made, as well as the absolute values for these measurements (which would be easily achievable through the use of a standard curve of ATP), these data are uninterpretable. Furthermore, it seems unlikely that yeast would be able to accommodate shifts of ATP levels that span an order of magnitude without dire cellular consequences, particularly during rapamycin treatment.

We appreciate the valuable feedback from the reviewer regarding the importance of providing detailed information on yeast growth conditions for interpreting ATP data. In response to this suggestion, we have enhanced the figure legends associated with the relevant figures to include a comprehensive description of the yeast growth conditions. This now specifies the age of the culture, culture media composition, and other pertinent parameters.

In addressing the concern raised about the rapamycin-induced ATP increase, we have carefully re-examined our experimental procedures. We performed additional experiments and confirmed the consistency of our findings in logarithmic-treated cultures. The results remain in alignment with our initial observations, reinforcing the reliability and reproducibility of our data.

(3) As stated above, the inference of mitochondrial function from cellular ATP levels, cellular ROS levels, and gene expression of a handful of nuclear-encoded genes is not sound. The authors should include further experimentation as evidence of mitochondrial functionality, such as respirometry or metabolic flux experiments.

Thank you for your constructive feedback on our manuscript. We appreciate your careful consideration of our work. In response to your concerns regarding the indirect nature of our mitochondrial function assays, we have implemented the following changes: We have incorporated additional assays to provide a more direct insight into mitochondrial biology. Specifically, we performed relative mitochondria content analysis (S3F) and quantified ROS levels under fermentative to stationary phase conditions (S3G). These assays offer a more direct and comprehensive assessment of mitochondrial function. Furthermore, we conducted experiments in respiratory glycerol medium (S3H) to complement our previous findings.

To further support our claims, we investigated the resistance of ybr238c∆ cells to H2O2 toxicity. Our results demonstrate that these cells exhibit increased resistance compared to wild-type cells. This additional evidence strengthens the link between mitochondrial function and cellular response to oxidative stress.

We believe these adjustments address your concerns and significantly enhance the robustness of our study. We hope you find these modifications satisfactory. We are grateful for your valuable input, which has undoubtedly improved the clarity and reliability of our findings.

(4) Multiple gene expression analyses are performed on n=2 measurements, and this should be bolstered by further replicates. Many bar graphs do not have accompanying statistics; these should be added. Some statistical tests are performed across inappropriate comparisons, such as Figure 3G, in which expression levels of mitochondrial genes in both deletion and overexpression strains should be compared to a wild-type control rather than to each other.

Thank you for your thorough review and constructive feedback on our manuscript. We appreciate your careful examination of our work. In response to your comments, we have made the following revisions to address your concerns: The multiple gene expression analysis in our study focused specifically on ETC genes. It is important to note that ETC genes themselves represent multiple replicates within the ybr238c deletion and overexpression cells, as illustrated in Figures 4D, 4G, and 6B.

We acknowledge and appreciate your observation regarding Figure 3G. To address this concern, we have revised the statistical comparisons. The expression levels of mitochondrial genes in the overexpression strain are now appropriately compared to a wild-type control. This correction has been applied in the figure that correctly corresponds to text in the manuscript.

(5) Figure 2B is uninterpretable as it stands, as most gene symbols are obscured.

We appreciate the reviewer's attention to Figure 2B and the feedback provided. Regarding the gene labels in Figure 2B, we would like to clarify that these labels are default outputs from the Metascape analysis, and unfortunately, repositioning them within the current figure layout isn't feasible without compromising the integrity of the information.

However, we have taken the reviewer's concern seriously and have made efforts to address the interpretability issue. To provide readers with access to the full set of genes for functional enrichment analysis and the MCODE complex, we have included this information in Additional File 3. The figure legends have been updated accordingly to guide readers to refer to Additional File 3 for a more detailed examination of the gene symbols and their annotations.

We hope that this solution addresses the concern raised by the reviewer.

(6) The conclusions to be drawn from Figure 3A are not clear, and this figure is cited only once in the text along with two other figures (page 8).

Thank you for your valuable feedback. We have carefully considered your comments and made revisions to improve the clarity of the conclusions drawn from Figure 3A.

(7) Figure 6K reports a range of 100-200% cell survival - how does a cell have 200% survival? Isn't survival binary (i.e., you survive or you are dead)? Perhaps this is meant to be relative to another condition; this should be more clearly stated in the figure, or the axis should be normalized to a maximum of 100% survival.

Thank you for your guidance and valuable feedback. Based on your recommendation, we have made significant changes to Figure 6K in the revised manuscript. Specifically, we replaced the heatmap with a bar graph to enhance clarity. Additionally, we would like to highlight that cell survival of combined treated cells is measured relative to the control treatment, which is considered 100% survival. This aims to provide a more accurate and comprehensible representation of the data. We believe these modifications contribute to a clearer presentation of our findings.

(8) The authors state that "TORC1 inhibition in yeast and human cells with mitochondrial dysfunction suppresses their accelerated aging." No studies of aging were done in human cells; survival in response to mitochondrial toxins does not reveal aging phenotypes. To state such is a substantial overstatement and should be amended to perhaps "cellular survival" rather than directly linked to aging.

We appreciate the careful review of our manuscript and the constructive feedback provided by the reviewer. In response to the concern raised regarding the statement about TORC1 inhibition and accelerated aging in human cells, we have revised the relevant passage as follows: "In turn, TORC1 inhibition in yeast and human cells with mitochondrial dysfunction enhances their cellular survival." We believe that this modification accurately reflects the outcomes of our experiments and addresses the concern raised by the reviewer. We would like to express our gratitude for the valuable feedback, which has contributed to the improvement of our manuscript. Thank you for your thoughtful consideration.

Reviewer #3 (Recommendations For The Authors):

Thank you for your detailed review and valuable recommendations. We have carefully addressed each of your comments in the revised manuscript. The specific changes made include:

The authors should have attempted to fully characterize the RLS and CLS phenotype of strains lacking the YBR238C and RMD9 gene, the single most important gene identified in this study. Before further characterization, its association with aging must be tested to replicate findings from the literature. Although Figure 3 shows partially characterized CLS in SC medium, different media conditions could be tested, and the full spectrum of CLS lifespan curves should be represented. RLS phenotypes of these cells were not analyzed throughout the study.

We appreciate the suggestion to investigate the role of YBR238C in both Replicative Lifespan (RLS) and Chronological Lifespan (CLS). However, it's essential to note that the involvement of YBR238C in the RLS phenotype has been previously documented in four studies (references 7, 39, 40, and 41). Considering the established literature on this matter, we chose not to duplicate these efforts in our study.

Our primary focus was on confirming the role of YBR238C in the chronological lifespan (CLS) phenotype, as indicated by a genome-wide study (reference 43). Accordingly, we also conducted an analysis of the role of RMD9 in CLS. The methods and figure legends explicitly state that CLS experiments for prototrophic CEN.PK113-7D strains were conducted in synthetic defined (SD) medium containing 6.7 g/L yeast nitrogen base with ammonium sulfate without amino acids and 2% glucose. For auxotrophic BY4743 strains, SD medium was supplemented with histidine (40 mg/L), leucine (160 mg/L), and uracil (40 mg/L).

It is important to clarify that SC medium was not used for CLS analysis. Instead, we employed SD medium, recommended for CLS analysis (reference 15; PMID: 22768836). The CLS experiments were conducted using three different methods, providing a comprehensive representation of the entire CLS lifespan (Figures 1C, 1D, 1E, and 1F).

While we did not present the Replicative Lifespan (RLS) phenotype explicitly, we performed experiments such as mitochondrial activity and ROS production under both CLS and RLS conditions. These additional analyses contribute valuable insights into the broader implications of YBR238C and RMD9 on cellular function.

We believe that these clarifications and the inclusion of additional experimental details enhance the robustness and validity of our findings. We hope these explanations address the concerns raised by the reviewer and contribute to the overall improvement of our manuscript.

In addition, authors include RNAseq data from Rapamycin-treated cells to identify differentially expressed genes. Notably, genes with decreased expression were used to compare KO strains' lifespan phenotype. Additional RNAseq analyses were performed on individual KO cells. The methodology section needs to be better written with information on which media and metabolic state that these cells are collected after treatment with rapamycin. If the cells are collected during logarithmic growth, the data can be compared with RLS aging gene sets only. A separate experiment has to be performed on stationary cells (respiratory) to collect RNAseq data after rapamycin treatment, then can be compared to the CLS aging gene set.

Thank you for your insightful comments and considerations regarding our methodology for obtaining Rapamycin response genes (RRGs). We appreciate the opportunity to address your concerns and provide further clarification on our experimental approach.

As mentioned in our manuscript, we obtained RRGs by treating logarithmic cells with 50 nM Rapamycin for 1 hour, and the details have been included in supplementary Figure S1C legends. Our primary objective was to compare these RRGs with agingassociated genes that modulate both Replicative Lifespan (RLS) and Chronological Lifespan (CLS). We acknowledge the significance of this comparison and believe that our approach, treating logarithmic cells, is suitable for achieving this goal.

It is important to note that the use of a higher concentration of Rapamycin for treatment renders the cells less efficient in terms of growth, resulting in a very low optical density (OD) at 72 hours, as illustrated in Figure 6H. Unfortunately, due to this limitation in growth efficiency, obtaining Rapamycin response genes at the stationary phase was not feasible in our experimental setup.

As the experimental conditions vary among the reports and the gene expression signature significantly changes under different metabolic conditions, the media condition that samples are collected for RNAseq analyses should match the media condition that the lifespans of those KO strains are tested. However, more information needs to be detailed on these methodologies. For example, the transcriptomic signature of the YBR238C KO strain should be done under both fermentative and respiratory conditions to understand the true gene expression signature associated with CLS and RLS. Throughout the manuscript, these two metabolic conditions and associated lifespan types (CLS vs. RLS) are not differentiated and treated as the same, probably causing the biggest confounding effect that resulted in the identification of a single yeast-specific gene.

We obtained the transcriptomic signature of the YBR238C KO strain from logarithmic phase cultures. This consistency was maintained to align with the Rapamycin Response Genes (RRGs) obtained from logarithmic cells treated with rapamycin. Detailed methodology and metabolic status information is provided in the method section and relevant figure legends.

To broaden the scope of our study, we conducted analyses on various phenotypes, including mitochondrial activity and oxidative stress, under both logarithmic phase (relevant to Replicative Lifespan, RLS) and stationary phase (relevant to Chronological Lifespan, CLS). We have now explicitly indicated the logarithmic phase/stationary phase conditions in the figure legends of the manuscript, specifying their relevance to RLS or CLS.

Results from these additional experiments have been incorporated into the revised manuscript as supplementary figures (S3E-S3I). We believe that these clarifications and the inclusion of additional experimental details enhance the robustness and validity of our findings. We trust that these explanations effectively address the concerns raised by the reviewer and contribute to the overall improvement of our manuscript.

YBR238C gene KO effect on mitochondrial function missing comprehensive characterization. Whether the improved mito function caused by increased mtDNA copy number and/or increased mitochondrial number could be easily tested by analyzing normalizing RNAseq reads from mtDNA genes to reads from nucDNA genes. Data could be further combined with western blot specific to mito membrane proteins to analyze mito copy number.

Thank you for your insightful comments and suggestions. Following your recommendation, we conducted an assessment of relative mitochondrial content (see Figure S3F) and observed significantly higher mtDNA content in the ybr238c∆ compared to the wild type (see Figure S3F). Additionally, we have incorporated the methodology for mitochondrial DNA copy number analysis in the methods section.

The two paralogous gene interaction is an interesting observation. However, in yeast, it is known that deletion of one of the paralogous genes causes copy number amplification of the certain chromosome that the other paralogous gene is located, causing aneuploid chromosome. Many of the observed phenotypes can be associated with increased chromosome copy number and should be carefully tested. However, the authors did not consider this important point. Simply, using RNA seq data normalized read/per chromosome could be plotted to analyze the karyotype of YBR238C and RMD9 KO cells.

We appreciate your thoughtful consideration of our work and the suggestion to investigate chromosome copy number variations. While we did not directly test the chromosome copy, we want to highlight that our study extensively explores the impact of YBR238C on cellular lifespan through an RMD9-dependent mechanism (Figure 5). Deletion of YBR238C increases, whereas overexpression of YBR238C decreases the expression of its paralog, RMD9 (Figure 5F). Furthermore, this phenotype is associated with the lifespan of YBR238C-deleted and overexpressed cells. In our study, we have thoroughly investigated this aspect.

Version 2

Abstract

Abstract

Uncovering the regulators of cellular aging will unravel the complexity of aging biology and identify potential therapeutic interventions to delay the onset and progress of chronic, aging-related diseases. In this work, we systematically compared gene sets involved in regulating the lifespan of Saccharomyces cerevisiae (a powerful model organism to study the cellular aging of humans) and those with expression changes under rapamycin treatment. Among the functionally uncharacterized genes in the overlap set, YBR238C stood out as the only one downregulated by rapamycin and with an increased chronological and replicative lifespan upon deletion. We show that YBR238C and its paralogue RMD9 oppositely affect mitochondria and aging. YBR238C deletion increases the cellular lifespan by enhancing mitochondrial function. Its overexpression accelerates cellular aging via mitochondrial dysfunction. We find that the phenotypic effect of YBR238C is largely explained by HAP4- and RMD9-dependent mechanisms. Further, we find that genetic or chemical-based induction of mitochondrial dysfunction increases TORC1 (Target of Rapamycin Complex 1) activity that, subsequently, accelerates cellular aging. Notably, TORC1 inhibition by rapamycin (or deletion of YBR238C) improves the shortened lifespan under these mitochondrial dysfunction conditions in yeast and human cells. The growth of mutant cells (a proxy of TORC1 activity) with enhanced mitochondrial function is sensitive to rapamycin whereas the growth of defective mitochondrial mutants is largely resistant to rapamycin compared to wild type. Our findings demonstrate a feedback loop between TORC1 and mitochondria (the TORC1-MItochondria-TORC1 (TOMITO) signaling process) that regulates cellular aging processes. Hereby, YBR238C is an effector of TORC1 modulating mitochondrial function.

Introduction

Healthy aging is crucially determined by cellular functions, and their defective status is associated with premature dysfunction and/or depletion of critical cell populations and various aging-associated pathologies such as neurodegenerative diseases, cancer, cardiovascular disorders, diabetes, sarcopenia and maculopathy 1–5. Advancements in biomedical research including genome-wide screening in different aging model organisms, identified several biological pathways that support the unprecedented progress in understanding overlapping profiles between aged cells and different chronic aging-associated diseases 6–9. In principle, uncovering the molecular mechanisms that drive cellular aging identifies potential drug targets and can fuel the development of therapeutics to delay aging and increase healthspan 10–13.

Given the evolutionary conservation of many aging-related pathways, yeast is one of the aging model organisms that have been extensively used to study the biology of human cellular aging by analyzing chronological lifespan (CLS) and replicative lifespan (RLS) under various conditions 14–17. The CLS is the duration of time that a non-dividing cell is viable. This is a cellular aging model for post-mitotic human cells such as neurons and muscle cells. The RLS defines as the number of times a mother cell divides to form daughter cells. Such experiments provide a replicative human aging model for mitotic cells such as stem cells. Genome-wide or individual gene deletion strains’ screening identified thousands of genes affecting cellular lifespan. These lists are a rich resource to identify unique genetic regulators, functional networks, and interactions of aging hallmarks relevant to cellular lifespan 1,4,6,10–13. At the same, these lists are rich in genes of unknown function, a class of genes that, unfortunately, got increasingly ignored by the attention of research teams 18–20.

We systematically examined the available genetic data on aging and lifespan for budding yeast, Saccharomyces cerevisiae, from various sources. Our bioinformatics analyses revealed consensus lists of genes regulating yeast CLS and RLS. These gene sets were compared with lists of genes the expression of which changed under TORC1 inhibition with rapamycin. When we turned our attention to the functionally uncharacterized genes involved, we found YBR238C the only one among the latter that increases both CLS and RLS upon deletion and that is downregulated by rapamycin. Transcriptomics and biochemical experiments revealed that YBR238C negatively regulates mitochondrial function, largely via HAP4- and RMD9-dependent mechanisms, and thereby affects cellular lifespan. Surprisingly, YBR238C and its paralogue RMD9 oppositely influence mitochondrial function and cellular aging.

Our chemical genetics and metabolic analyses unravel a feedback loop of the interaction of TORC1 with mitochondria that affect cellular aging. YBR238C is an effector of TORC1 that modulates mitochondrial function. We also show that mitochondrial dysfunction induces TORC1 activity enhancing cellular aging. In turn, TORC1 inhibition in yeast and human cells with mitochondrial dysfunction enhances their cellular survival.

Results

Genome-wide survey of genes affecting cellular lifespan in yeast in accordance with literature and public databases

Lists of scientific literature instances mentioning a yeast gene as affecting lifespan were downloaded from the databases SGD (Saccharomyces cerevisiae genome database) 21, and GenAge 22. In most cases, the experiments referred to are gene deletion phenotype studies. After processing the files for the mentioning of ‘increase/decrease’ of ‘chronological lifespan’ (CLS) and/or ‘replicative lifespan’ (RLS) as well as the suppression of gene duplicates, we found 2399 entries with distinct yeast genes in 15 categories reported to increase/decrease CLS and/or RLS under various conditions (Table 1). We collectively call them Aging Associated Genes (AAGs). Notably, about one-third of the total yeast genome belongs to that category. Downloaded files (as of 8th November 2022), description of the processing details, and all the resulting gene lists are available in ‘Additional File 1’.

Whereas most of the genes (1610, 67%) have been mentioned for just one of the four scenarios (“CLS increase”, “CLS decrease”, “RLS increase”, and “RLS decrease”), the remaining genes have been described for several alternative and/or even opposite/conflicting outcomes. 19 genes are brought up in context with all four scenarios. As the experimental conditions varied among the reports and the gene networks are complex, this is not necessarily a contradiction. Yet (see below) some of the cases are actually annotation errors at the database level.

We explored the potential enrichment of gene ontology (GO) terms among the genes involved in the 15 categories with the help of DAVID WWW server 23. The most significant result is the enrichment of ontology terms related to mitochondrial function among the genes annotated with the single qualifier “CLS decrease”. The top mitochondrial term cluster has an enrichment score 18.2 (gene-wise P-values and Benjamini-Hochberg values all below 6.e-13); a second one related to mitochondrial translation has the score 4.7. Enrichment of mitochondrial ontology terms has also been observed for “CLS decrease” in combination with “CLS increase” or “RLS increase/decrease”.

Every other signal from the ontology is much weaker. Term enrichment (with scores near 2 or better) related to autophagy and vesicular transport pop up for CLS-annotated genes whereas terms connected with translation, DNA repairs, telomeres, protein degradation and signaling are observed with RLS-tagged gene lists.

We also explored the presence of uncharacterized or severely under-characterized genes in the 15 categories of AAGs (Table 1). In total, 944 genes are annotated in SGD as coding for a protein of unknown function. Additionally, we considered genes without dedicated gene name (only with six-letter locus tag) as dramatically under-characterized. Comparison of this combined list with those in the 15 categories reveals 88 severely understudied AAGs that are candidates for enhanced attention from the scientific community. Thus, functionally insufficiently characterized genes have a great role in cellular aging-related processes.

Rapamycin response genes overlap with the AAGs

Nutrient sensing dysregulation is one of the aging hallmarks 10,11. The conserved protein complex Target of Rapamycin Complex 1 (TORC1) senses nutrients such as amino acids and glucose and links metabolism with cellular growth and proliferation 24–27. TORC1 positively regulates aging, and its inhibition increases lifespan in various eukaryotic organisms including yeast and mammals 12,25,26,28,29. The drug rapamycin, initially discovered as an antifungal natural product produced by Streptomyces hygroscopicus, inhibits TORC1 and increases lifespan (Figures S1A and S1B) 30,31.

To explore the connection between nutrient signaling and cellular aging we mapped the TORC1 regulated genes with AAGs. We first identified rapamycin response genes (RRGs) by transcriptomics analysis (RNA-Seq for yeast S. cerevisiae BY4743 cells treated with rapamycin and DMSO control; Figure S1C). Relevant measurement results, lists of 2365 RRGs and supplementary methodical comments are available in ‘Additional File 2’.

As overlap of two RNA-Seq data analysis methods (see Methods), we identified 1155 rapamycin upregulate genes (RUG) and 1210 rapamycin downregulated genes (RDG) (Figures S1D-S1F; Additional File 2). RNA-Seq results were confirmed for some genes by qRT-PCR (Figure S2A). We also checked the differential gene expressions in prototrophic yeast S. cerevisiae strain CEN.PK and found similar results as with the BY4743 strain (Figure S2B). To note, our transcriptomics analyses are, as a trend, consistent with previous RNA-Seq studies carried out under partially different experimental conditions 32,33.

We mapped the RRGs (RUG and RDG) with AAGs. Among the 2399 AAGs names, 397 and 433 (in total 830) re-occur in the lists of RUG and RDG, respectively. Thus, the overlap with AAGs is nearly 35%. I.e., TORC1 controls about one third of the AAGs. Table 1 shows how many AAGs are up- and down-regulated by rapamycin treatment for each of the 15 categories with regard to CLS/RLS increase/decrease. Notably, the order of magnitude for the respective numbers of RUG and RDG is the same for all 15 studied subgroups.

Since rapamycin increases the lifespan by an inhibitory effect on TORC1 activity, it appears most interesting to focus on AAGs with a deletion phenotype of increased CLS/RLS and being downregulated by rapamycin application. A manual analysis of these gene lists reveals that, among the uncharacterized or severely understudied AAGs, there is a single one known to increase both CLS and RLS upon deletion and being downregulated by rapamycin treatment. This gene is YBR238C.

Uncharacterized genes mapping unravels the role of YBR238C in cellular aging

Not much is known about YBR238C besides its effect on lifespan (to increase CLS and RLS), its mitochondrial localization 34, its transcriptional up-regulation by TORC1 and the existence of the paralogue RMD9. The encoded protein has 731 amino acid residues. Sequence architecture analysis with the ANNOTATOR 35 reveals an intrinsically unstructured region over the first ca. 130 residues (first, a long polar but uncharged run followed by a histidine/asparagine-rich region beginning with position 83) and a pentatricopetide repeat region (residues 130-675, for example, due to a HHpred 36 hit to structure 7A9X chain A with E-value 2.e-56). Given the sequence homology, we hypothesize that the protein encoded by YBR238C is involved in RNA binding as its paralogue RMD9 with a similar globular segment37.

To note, YBR238C carries the conflicting annotations in SGD for increased and decreased RLS upon deletion. Unfortunately, there is not a single direct report about YBR238C listed in the scientific literature at the time of writing. There are a few genome-wide deletion strain studies that identified YBR238C as one of the gene that increases CLS 38 and RLS 7,39–41. However, the SGD database wrongly documented one of the latter studies as evidence for a decreased RLS phenotype of YBR238C 41. Thus, this examination allows us to claim YBR238C as the only uncharacterized rapamycin downregulated gene causing CLS and RLS increase upon deletion.

First, we confirmed that YBR238C is indeed a rapamycin response gene by qRT-PCR expression analysis in both yeast backgrounds BY4743 and CEN.PK (Figures 1A and 1B). Given that only a single genome-wide deletion strain study identified YBR238C as a gene that enhances CLS 38, we further tested the role of YBR238C in CLS of yeast. CLS was analysed in BY4743 and CEN.PK strains using three different outgrowth survival methods (see detail in methods section). Cell survival of wild type and ybr238cΔ strains was analysed at various age time points. We found higher CLS of ybr238cΔ cells compared to wild type cells (Figures 1C-1F). Together, these results confirmed that rapamycin inhibits the expression of YBR238C, and deletion of this gene indeed increases the cellular lifespan.

YBR238C deletion increases the cellular lifespan

(A and B) Expression analysis of YBR238C gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (A) and CEN.PK113-7D (B). qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 (A) and only the synthetic defined medium for CEN.PK113-7D (B). Data are represented as means ± SD (n=4). ****P < 0.0001 based on two-sided Student’s t-test.

(C and D) Chronological lifespan (CLS) of the wild type and ybr238cΔ mutant was assessed in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 (C) and only the synthetic defined medium for CEN.PK113-7D (D) strains in 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=3) (C) and (n=4) (D). **P < 0.01, ***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Šídák’s multiple comparisons test.

(E and F) CLS of the wild type and ybr238cΔ mutant was performed in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 and only the synthetic defined medium for CEN.PK113-7D strains using flasks. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (E) and YPD medium in the 96-well plate (F). The serial outgrowth of aged cells on agar medium was imaged (E) and quantified the survival relative to outgrowth of day 2 (F). A representative of two experiments for (E) and (F) is shown.

Transcriptomics analysis reveals the longevity gene expression signatures of ybr238cΔ mutants

The transcriptome of the long-lived ybr238cΔ mutant was compared with that of the wild type (Figure S3A). Applying standard significance criteria, we found 326 genes up- and 61 genes down-regulated in the ybr238cΔ mutant compared to wild type (Figure S3B; Additional File 3). Thus, the transcriptome of the ybr238cΔ mutant is very distinct from that of the wild type.

Notably, we see several major metabolic changes. For the ybr238cΔ mutant, genes in mitochondrial metabolic processes such as oxidative phosphorylation and aerobic respiration show predominant enrichment (Figures 2A and 2B; Additional File 3). Since YBR238C expression is regulated by TORC1 (Figures 1A and 1B), it is not surprising that we observe similarities in the profiles of upregulated DEGs for the ybr238cΔ mutant and for the case of treating the wild type with rapamycin (Figures S3C and S3D).

Longevity signatures of ybr238cΔ mutant

(A and B) Functional enrichment analysis of upregulated genes in yeast Saccharomyces cerevisiae ybr238cΔ mutant compared to wild type (BY4743). Bar plots showing the enriched biological process (A) and identified MCODE complexes based on the top-three ontology enriched terms (B). Full set of genes for functional enrichment analysis (A) and the MCODE complexes (B) are available in Additional File 3.

(C and D) ATP analysis of wild type and ybr238cΔ mutant for BY4743 (C) and CEN.PK113-7D (D). Quantification was performed using ATP extracted from 72-hour stationary cultures grown in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 and only the synthetic defined medium for CEN.PK113-7D strains. Data are represented as means ± SD (n=6). **P < 0.01, and ****P < 0.0001 based on two-sided Student’s t-test.

(E and F) Expression analysis of MSN4 gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (E) and CEN.PK113-7D (F). qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 and only the synthetic defined medium for CEN.PK113-7D strains. Data are represented as means ± SD (n=2). **P < 0.01 based on two-sided Student’s t-test.

(G and H) ROS analysis of wild type and ybr238cΔ mutant for BY4743 (G) and CEN.PK113-7D (H). Quantification was performed for logarithmic-phase cultures grown in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 and only the synthetic defined medium for CEN.PK113-7D strains. Data are represented as means ± SD (n=4). ***P < 0.001, and ****P < 0.0001 based on two-sided Student’s t-test.

Next, we experimentally tested whether the transcriptome longevity signatures is associated with enhanced mitochondrial metabolism, whether the cellular energy level has gone up and cellular stress responses are induced with a switch to oxidative metabolism 42,43. Indeed, our metabolic analysis revealed an increased ATP level in ybr238cΔ mutants compared to wild type cells (Figures 2C and 2D; S3E). ATP generation through the oxidative phosphorylation (OXPHOS) system is regulated by mitochondrial DNA (mtDNA) 44. We observed a higher mtDNA copy number in the ybr238cΔ mutant compared to wild type cells (Figure S3F). Hence, the ybr238cΔ mutation rewires the cellular metabolism to promote resource-saving ways of energy production as the up-regulated expression of OXPHOS machinery subunits truly boosts ATP synthesis in ybr238cΔ mutant cells.

The enhanced mitochondrial function in ybr238cΔ mutants does also improve the protection against reactive oxygen species (ROS). Among the 150 TFs that control the upregulated DEG of the ybr238cΔ mutant (Additional File 3), 13 TFs are significantly overrepresented within the upregulated DEGs (Additional File 3). Importantly, we found that the stress response controlling transcription factor MSN4 is upregulated in ybr238cΔ mutants (Figures 2E and 2F; S4A-S4B) 15,45. Concomitantly, we found less ROS level in ybr238cΔ mutant compared to wild type cells (Figures 2G and 2H; S3G and S3H) and is resistant to H2O2 induced oxidative stress toxicity (Figure S3I).

Notably, the transcription factor regulated activation of stress response pathways (thioredoxins, molecular chaperones, etc.) 46,47 and the switch from fermentation to respiration are associated with delayed cellular aging 42,43,48,49. Our results show that the ybr238cΔ mutation triggers oxidative metabolism and stress protective machineries activation and, as a result increases CLS. Collectively, our findings reveal that YBR238C is a TORC1 regulated gene involved in mitochondrial function coupled with cellular aging. Therefore, we refer to YBR238C as AAG1: Aging Associated Gene 1.

YBR238C negatively regulates mitochondrial function via HAP4-dependent and -independent mechanisms

HAP4 is a transcription factor that controls the expression of mitochondrial electron transport chain components including OXPHOS genes 45. HAP4 activity has been shown to increase lifespan by enhancing the mitochondrial respiration in cells 45. Intriguingly, HAP4 is among the 13 TFs overrepresented among the upregulated DEGs in ybr238cΔ mutant and control mitochondrial genes (Figure 3A; Additional File 3). We validated the upregulation of HAP4 gene expression in the ybr238cΔ mutant under both yeast backgrounds BY4743 and CEN.PK (Figures 3B and 3C). Consistent with these findings, transcription profile of mitochondrial genes in the ybr238cΔ mutant is opposite to that of the hap4Δ mutant (Figure 3D). For example, ETC complexes I – V genes’ expression is increased in ybr238cΔ, however, it is decreased by HAP4 deletion (Figure 3D).

YBR238C affects cellular aging via HAP4-dependent and independent mechanisms

(A) The HAP4 regulon within the upregulated DEGs of ybr238cΔ mutant. See also Additional File 3.

(B and C) Expression analysis of HAP4 gene by qRT-PCR in yeast Saccharomyces cerevisiae genetic backgrounds BY4743 (B) and CEN.PK113-7D (C). qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium supplemented with auxotrophic amino acids for BY4743 and only the synthetic defined medium for CEN.PK113-7D strains. Data are represented as means ± SD (n=2). **P < 0.01 based on two-sided Student’s t-test.

(D) Expression analysis of mitochondrial ETC genes by qRT-PCR in wild type, ybr238cΔ, hap4Δ, and ybr238cΔ hap4Δ, strains of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D. qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium. Data are represented as means ± SD (n=2). Comparing data between ybr238cΔ and ybr238cΔ hap4Δ. ***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Šídák’s multiple comparisons test.

(E, F and H) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=6). *P < 0.05, ***P < 0.001, ****P < 0.0001 and ns: non-significant. Ordinary one-way ANOVA followed by Tukey’s multiple comparisons test (E and F). Two-way ANOVA followed by Šídák’s multiple comparisons test (H).

(G) Expression analysis of HAP4 and ETC genes by qRT-PCR in wild type and YBR238C overexpression (YBR238C-OE) strains of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D (C). qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium. Data are represented as means ± SD (n=2). *P < 0.05, **P < 0.01, and ***P < 0.001 based on Two-way ANOVA followed by Šídák’s multiple comparisons test.

(I) Model representing regulation of cellular aging by YBR238C via HAP4-dependent and independent mechanisms.

We found that deletion of HAP4 moderately decreases CLS at the background of the ybr238cΔ mutant (Figure 3E). To confirm that the decrease in lifespan is through the HAP4 pathway, we examined the expression of mitochondrial respiratory genes. We found that HAP4 deletion significantly decrease the ETC complex I – V genes’ expression in ybr238cΔ mutant (Figure 3D). To note, HAP4 deletion at the wild-type background decreases the cellular lifespan even more (Figure 3E), which is consistent with more dramatic reduction of ETC complexes I – V genes’ expression (Figure 3D). Taken together these data suggests that YBR238C negatively regulates the HAP4 activity. HAP4-upregulated, increased mitochondrial function contributes to the prolonged lifespan of ybr238cΔ cells.

YBR238C gene deletion rescues some loss of lifespan of hap4Δ cells (Figure 3F). The effect is especially pronounced until day 8 when wild-type cells are essentially 100% surviving. Yet, complete epistasis of phenotypes is not achieved. This observation parallels the above findings that HAP4 deletion in ybr238cΔ mutant does not fully recover the mitochondrial ETC complexes I – V genes’ expression and CLS at day 9 compared to the wild type background (Figures 3D and 3E). Together, these results indicate that YBR238C affects cellular lifespan via HAP4-dependent and -independent mechanisms.

To confirm the existence of HAP4-independent mechanisms, we examined the lifespan and transcriptome under conditions of YBR238C overexpression (YBR238C-OE). As expected, YBR238C-OE decreases the expression of HAP4, of mitochondrial ETC complexes I – V genes (Figure 3G) as well as the CLS (Figure 3H). Strikingly, the lifespan of YBR238C-OE cells was shorter than for hap4Δ cells (Figure 3H). Thus, a HAP4-independent mechanism does exist through which YBR238C also affects cellular aging (Figure 3I).

The YBR238C paralogue RMD9 deletion decreases the lifespan of cells

In yeast S. cerevisiae, YBR238C has a paralogue RMD9 that shares ∼45% amino acid identity 50. Given that YBR238C activity is tightly coupled with the lifespan, we investigated the role of its paralogue RMD9 in cellular aging. Surprisingly, we found that RMD9 deletion has an effect opposite to YBR238C deletion and it shortens CLS (Figures 4A-4C; S5A-S5C).

RMD9 deletion decreases the cellular lifespan

(A and E) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=12) (A) and (n=6) (B). ****P < 0.0001 and ns: non-significant. Ordinary one-way ANOVA followed by Tukey’s multiple comparisons test (A). Two-way ANOVA followed by Dunnett’s multiple comparisons test (E).

(B, C, F and G) CLS of yeast strains was performed in synthetic defined medium using flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (B and F) and YPD medium in the 96-well plate (C and G). The serial outgrowth of aged cells on agar medium was imaged (B and F) and quantified the survival relative to outgrowth of day 1 (C and G). A representative of two experiments for B, C, F and G is shown.

(D) Serial dilution growth assays of the yeast strains on fermentative glucose medium (YPD) and respiratory glycerol medium (YPG).

(H) ATP level in yeast strains. Quantification was performed using ATP extracted from 72-hour stationary cultures grown in synthetic defined medium. Data are represented as means ± SD (n=3). *P < 0.05, and ***P < 0.001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test.

See also Figure S5.

RMD9 was initially isolated during a search for genes required for meiotic nuclear division 51. Subsequently, its role was uncovered in controlling the expression of mitochondrial metabolism genes by stabilizing and processing mitochondrial mRNAs 37,50,52. We asked whether RMD9 deletion induces mitochondrial dysfunction that, therefore, causes accelerated cellular aging. We first tested the mitochondrial activity by allowing mutant cells to grow under respiratory conditions 53–55. We found that rmd9Δ mutants could perform on glucose medium, but they failed to grow on glycerol as a carbon source (known to require functional mitochondria; Figures 4D; S5D). Our results are in line with previously observed dysfunctional mitochondria in rmd9Δ mutants 37,50.

In control experiments, we found a similar growth defect phenotype on glycerol medium for cells deficient in PET100 and COX6 mitochondrial genes (Figures 4D; S5D) 55. CLS of pet100Δ and cox6Δ mutants were reduced compared to wild type (Figures 4E-4G; S5E-S5G). We also found lowered ATP- and higher ROS-levels in rmd9Δ, pet100Δ, cox6Δ mutants compared to wild type cells (Figures 4H and S5H).

In contrast, long lived ybr238cΔ mutants efficiently grow on respiratory medium with high ATP- and low ROS-levels (Figures 4D and 4H; S5D and S5H). So far, the results indicate that deletion of YBR238C potentiates the mitochondrial function that, in turn, leads to CLS increase. Thus, YBR238C and its paralogue RMD9 antagonistically affect mitochondrial function, CLS and cellular aging.

YBR238C affects the cellular lifespan via an RMD9-dependent mechanism

Previously, YBR238C deletion was shown to increase the CLS via HAP4-dependent and -independent mechanisms (Figure 3) and to rescue some loss of CLS of the hap4Δ mutant by a HAP4-independent mechanism (Figures 3E and 3F). Here, we ask whether YBR238C deletion suppresses the shortened lifespan of rmd9Δ mutants. We examined the lifespan of double deletion rmd9Δ ybr238cΔ mutant and compared it with rmd9Δ. We found that the deletion of YBR238C largely failed to recover the lifespan of rmd9Δ mutant to wild type cells; however, it partially prevented their early cell death (Figures 5A-5C). These results show that YBR238C deletion increases the cellular lifespan via RMD9-dependent mechanisms. Also, we found that the YBR238C deletion results in increased ATP- and reduced ROS-levels in the rmd9Δ mutant (Figures 5D and 5E).

YBR238C affects cellular aging via RMD9-dependent mechanism

(A, B, G and H) Chronological lifespan (CLS) of yeast Saccharomyces cerevisiae genetic background CEN.PK113-7D strains was performed in synthetic defined medium using flask. Outgrowth was performed for ten-fold serial diluted aged cells onto the YPD agar plate (A and G) and YPD medium in the 96-well plate (B and H). The serial outgrowth of aged cells on agar medium was imaged (A and G) and quantified the survival relative to outgrowth of day 2 (B and H). A representative of two experiments for A, B, G and H is shown.

(C and I) CLS of yeast strains was performed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=6) (C) and (n=12) (I). ****P < 0.0001 based on two-way ANOVA followed by Tukey’s multiple comparisons test (C) and Dunnett’s multiple comparisons test (I). ns: non-significant.

(D and J) ATP level in yeast strains. Quantification was performed using ATP extracted from 72-hour stationary cultures grown in synthetic defined medium. Data are represented as means ± SD (n=4). *P < 0.05, ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

(E and K) ROS level in yeast strains. Quantification was performed for logarithmic-phase cultures grown in synthetic defined medium. Data are represented as means ± SD (n=3). **P < 0.01, ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test. ns: non-significant.

(F) Expression analysis of RMD9 gene by qRT-PCR in yeast strains. qRT-PCR was performed using RNA extracted from logarithmic-phase cultures grown in synthetic defined medium. Data are represented as means ± SD (n=4). ***P < 0.001, and ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test.

(L) Effect of antimycin A (AMA) treatment on CLS of yeast strain was assessed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. A representative heatmap data of three experiments is shown.

(M) Model representing regulation of cellular aging by YBR238C via HAP4 and RMD9 dependent mechanisms.

Intriguingly, we find RMD9 expression upregulated in ybr238cΔ and downregulated in YBR238C-OE cells, respectively (Figure 5F). So, we asked whether RMD9 expression change is transcriptionally coupled in cellular lifespan phenotypes. Remarkably, RMD9 overexpression increases the lifespan of cells (Figures 5G-5I) as we would have predicted from the observed changes with the YBR238C deletion phenotype.

To know whether this CLS increase is contributed by enhanced mitochondrial function, we quantified the ATP level in wild type, YBR238C-OE, and RMD9-OE cells. ATP level is higher in RMD9-OE cells than wild type cells, a result in line with RMD9 positively regulating mitochondrial activity (Figure 5J). Consistent with the above findings, YBR238C overexpression decreases the ATP level (Figure 5J). Next, we assessed the oxidative stress and found that the ROS level in RMD9-OE cells was comparable to wild type. Yet, YBR238C overexpression increases the ROS level (Figure 5K).

It can be shown that the effects of YBR238C and RMD9 are at least partially realized via the mitochondrial ETC pathway. Antimycin A (AMA) is an inhibitor of the ETC complex III, decreasing ATP synthesis (Figure S6A) 56. Also, AMA treatment reduces the lifespan of cells (Figure S6B), confirming that mitochondrial energy supply is critical to delay cellular aging. Since YBR238C and RMD9 affect the ATP level, we tested the AMA effect on their deletion and overexpression strains. We found that ybr238cΔ and RMD9-OE cells with their enhanced mitochondrial function phenotype were largely resistant to AMA treatment (Figure 5L). AMA aggravates the cellular aging of mitochondrial defective YBR238C-OE cells (Figure 5L). Notably, mitochondrial defective rmd9Δ cells were not further affected by AMA treatment (Figure 5L).

Altogether, our findings reveal that YBR238C affects CLS and cellular aging via modulating mitochondrial function by mechanisms including HAP4- and RMD9-dependent pathways (Figure 5M).

YBR238C connects TORC1 signaling with modulating mitochondrial function and cellular aging

So far, we learned that YBR238C is regulated by TORC1 (Figures 1A and 1B) and it affects cellular lifespan by modulating mitochondrial function. Here, we wish establish the direct connection between the TORC1 signaling and mitochondrial activity. We treated cells with the TORC1 inhibitor rapamycin and found that, subsequently, the ATP level increases in the cells (Figures 6A; S6C). To test whether TORC1 affects mitochondrial function via YBR238C, we analyse the expression profile of mitochondrial ETC genes. As expected, rapamycin supplementation decreased the expression of YBR238C (Figures S6D and S6E) but induced the expression of ETC genes (Figure 6B). Remarkably, rapamycin-induced changes in the expression of ETC genes were largely unaffected in ybr238cΔ cells and reduced in YBR238C-OE cells (Figure 6B). Our results are consistent with the hypothesis that TORC1 regulates the mitochondrial ETC genes via YBR238C.

TORC1-MItochondria-TORC1 (TOMITO) signaling axis regulate cellular aging

(A) ATP analysis of logarithmic-phase wild type CEN.PK113-7D yeast cells treated with rapamycin for 1 hour in synthetic defined medium. Data are represented as means ± SD (n=3). ****P < 0.0001 based on ordinary one-way ANOVA followed by Dunnett’s multiple comparisons test. See also Figure S6C.

(B) Expression analysis of mitochondrial ETC genes by qRT-PCR. Analysis was performed using RNA extracted from logarithmic-phase wild type CEN.PK113-7D yeast cell treated with rapamycin for 1 hour in synthetic defined medium. Data are represented as means ± SD (n=2).

(C-G) Chronological lifespan (CLS) of yeast strains with indicated concentrations of rapamycin was performed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 2. Data are represented as means ± SD (n=3).

***P < 0.001, and ****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant.

(H) CLS of wild type yeast strain with indicated concentrations of rapamycin and antimycin A (10 µM) was performed in synthetic defined medium using 96-well plate. Aged cells survival was measured relative to the outgrowth of day 1. Data are represented as means ± SD (n=3).

****P < 0.0001 based on two-way ANOVA followed by Dunnett’s multiple comparisons test. ns: non-significant.

(I) ATP analysis of stationary-phase wild type CEN.PK113-7D yeast cells were incubated with rapamycin (10 nM) and antimycin A (10 µM) in synthetic defined medium. Data are represented as means ± SD (n=6). ****P < 0.0001 based on ordinary one-way ANOVA followed by Tukey’s multiple comparisons test. See also Figure S6H.

(J) Determination of human HEK293 cells survival incubated with indicated concentrations of antimycin A for 6 days. Untreated control is considered 100% cell survival. Cells survival was measured relative to the control. Data are represented as means ± SD (n=6).

(K) Cell survival analysis of human HEK293 cells incubated with indicated concentrations of antimycin A and rapamycin for 6 days. Antimycin A and rapamycin individual treated concentrations considered 100% cell survival control. Cells survival of combined treated cells was measured relative to the control. Data are represented as means ± SD (n=2). *P < 0.05, ****P < 0.0001, and ns: non-significant based on two-way ANOVA followed by Dunnett’s multiple comparisons test.

The strains ybr238cΔ and YBR238C-OE are associated with increased and decreased CLS, respectively. So, we ask whether TORC1 influences cellular aging via YBR238C by modulating mitochondrial function. We examined the effect of rapamycin supplementation on the lifespan of ybr238cΔ and YBR238C-OE cells. Strikingly, addition of rapamycin does not further increase CLS of ybr238cΔ cells (Figure 6C). Additionally, anti-aging effect of rapamycin is significantly reduced in YBR238C-OE cells (Figure 6C). These findings are consistent with the rapamycin effect on transcriptomics profiles of ETC genes in ybr238cΔ and YBR238C-OE cells (Figure 6B). Taken together, these results reveal that YBR238C is a downstream effector of TORC1 signaling, connecting mitochondrial function for the regulation of cellular aging.

Mitochondrial dysfunction induces the TORC1 activity that causes accelerated cellular aging

We showed that YBR238C-OE cells are associated with mitochondrial dysfunction and shortened lifespan (Figure 5). Apparently, the accelerated cellular aging phenotype is primarily due to compromised cellular energy level that affects the lifespan. Genetic (ybr238cΔ) and chemical (rapamycin) mediated enhancement of ATP level increases the lifespan of wild type cells validate this conclusion. Notably, the rapamycin supplementation significantly rescued the shortened lifespan of mitochondrial defective YBR238C-OE cells (Figures 6D; and S6F). This observation suggests that, as a reaction on sensing mitochondrial dysfunction, TORC1 is activated which, in turn, leads to accelerated cellular aging. Indeed, we found that cellular growth (a proxy for TORC1 activity) of mitochondria dysfunctional YBR238C-OE cells was resistant to rapamycin inhibition at concentrations that were effective for wild type cells (Figure S6G).

Further, we asked what the effect of TORC1 activity under enhanced mitochondrial function conditions would be. To test this, we assessed the growth of ybr238cΔ cells with rapamycin. We found that the cellular growth of ybr238cΔ cells was reduced by rapamycin compared to wild type cells (Figure S6G). Apparently, TORC1 activity is not signalled downstream under these conditions and rapamycin further aggravates this by TORC1 inhibition. This result suggests that YBR238C affects TORC1 activity via modulating mitochondrial function.

We found a similar pattern of rapamycin effect on growth of mutant cells with enhanced or damaged mitochondrial function. Growth of RMD9-OE cells having enhanced mitochondrial function was sensitive to rapamycin (Figure S6G). Likewise, growth of defective mitochondrial rmd9Δ, pet100Δ and cox6Δ mutants is resistant to rapamycin compared to wild type (Figure S6G). The YBR238C deletion reduced the rapamycin growth resistance phenotype of mitochondrial dysfunctional rmd9Δ (Figure S6G) and hap4Δ cells (Figure S6H), apparently by enabling enhanced mitochondrial function.

Taken together, we see that YBR238C plays a junction role in integrating mitochondrial function and TORC1 signaling.

TORC1 inhibition prevents accelerated cellular aging caused by mitochondrial dysfunction in yeast and human cells

Mitochondrial dysfunction increases the TORC1 activity and, subsequently, causes accelerating cellular aging. We asked whether inhibition of TORC1 activity could prevent accelerating cellular aging even under mitochondrial dysfunction conditions. Previously, we have already shown that rapamycin-mediated TORC1 inhibition partly rescued the shortened lifespan of mitochondrial defective YBR238C-OE cells (Figures 6D; and S6F). We examined other mitochondrial dysfunctional conditions to confirm that the suppressive effect of rapamycin is not only specific to YBR238C-OE. We tested defective mitochondrial mutants rmd9Δ, pet100Δ and cox6Δ. In all cases, rapamycin supplementation prevents the accelerated cellular aging (Figures 6E-6G).

Rapamycin supplementation also rescued the AMA-mediated accelerated cellular aging (Figure 6H). We also quantified the ATP level to confirm that the suppressive effect of rapamycin is specific to mitochondrial function. Rapamycin supplementation increases the ATP level of AMA-treated cells (Figures 6I; and S6I), demonstrating that TORC1 inhibition improves mitochondrial function and thus suppresses accelerated cellular aging.

We further verified the connection of TORC1 and mitochondrial function in human cells. First, we examined the effect of AMA on the survival of HEK293 cells. AMA treatment decreases the viability of HEK293 cells (Figure 6J). Cell survival reduction is due to the mitochondrial dysfunction as we see lower ATP levels in AMA treated cells compared to untreated control (Figure S6J). Subsequently, we examined the survival of AMA-treated HEK293 cells with or without rapamycin supplementation. Rapamycin supplementation suppresses AMA-mediated cellular death (Figure 6K). Further rapamycin supplementation increases the ATP level treated cells (Figure S6J).

Our findings convincingly support that TORC1 inhibition suppresses cellular aging associated with mitochondrial dysfunction across species.

Discussion

Identifying lifespan regulators, genetic networks, and connecting genetic nodes relevant to cellular aging provides functional insight into the complexity of aging biology 1,6,10 for the subsequent design of anti-aging interventions. Understanding the mechanism of aging will also require understanding the role of many genes of yet unknown function as YBR238C at the beginning of this work. Unfortunately, the group of uncharacterized genes coding proteins of unknown function has received dramatically decreasing attention during the past two decades 18–20.

In this study, we first summarized literature and genome database reports about genes affecting aging in the yeast model (mostly observed in gene deletion screens). We found that AAGs (Aging Associated Genes) make up about one-third of the total yeast genome (many of them are functionally uncharacterized) and can be classified into 15 categories for increase/decrease CLS and/or RLS under various conditions.

We compared the set of AAGs with the group of RRGs (Rapamycin Regulated Genes) and found that about one third of the AAGs is TORC1 activity regulated. Among the functionally uncharacterized genes in this overlap set, the gene YBR238C stands out as the only one that (1) is downregulated by rapamycin and (2) its deletion increases both CLS 38 and RLS 7,39–41. Thus, YBR238C is important among identified rapamycin-downregulated uncharacterized genes in AAG categories as it seems to rationally connect the rapamycin-induced TORC1 inhibition with the increase of both CLS and RLS.

We observed that the CLS of ybr238cΔ cells is higher than for wild type, regardless of various aging experimental conditions tested in this study. Transcriptional analysis of ybr238cΔ cells identified a longevity signature including enhanced gene expression related to mitochondrial energy metabolism and stress response genes 8,45,57–60. In contrast, YBR238C-OE cells display mitochondrial dysfunction leading to decreased lifespan. Together, these results reveal that TORC1 regulates the YBR238C expression which is linked to mitochondrial function and cellular lifespan.

The YBR238C paralogue RMD9 has been previously shown to affect mitochondrial function 37,50. Surprisingly, we observed an antagonism regarding the CLS phenotype of deletion/overexpression for YBR238C and RMD9. YBR238C overexpression and RMD9 deletion confer defective mitochondrial function associated with accelerated cellular aging and decrease lifespan of cells. In contrast, YBR238C deletion and RMD9 overexpression enhance the mitochondrial function that increase the cells’ longevity. Consistent with the identified negative regulatory role of YBR238C on mitochondrial function, deletion of this gene partially suppressed the accelerated aging of rmd9Δ cells and AMA-treated cells. We identified that YBR238C influences mitochondrial function and the CLS phenotype with contributions from HAP4- and RMD9-dependent mechanisms. As (i) one of RMD9’s molecular functions has been shown to stabilize certain (mitochondrial) mRNAs and (ii) YBR238C has a has a homologous pentatricopeptide repeat region, the two paralogues might differ in the sets of protected mRNAs with opposite outcome on cellular longevity.

While Kaeberlein et al. identified YBR238C as a top candidate for increasing replicative lifespan (RLS) in yeast through deletion 40, our study builds upon their work by investigating the mechanisms and the connection with TORC1 in chronological lifespan (CLS). Results of genetic interventions (YBR238C deletion and overexpression) and rapamycin treatment show that YBR238C is a downstream effector by which TORC1 modulates mitochondrial activity. Most importantly, we found that TORC1 inhibition increases the mitochondrial function via YBR238C and, thereby, extends cellular lifespan. TORC1 is involved in both CLS and RLS 15. Whether the TORC1 - YBR238C axis has similar or distinct mechanisms for CLS and RLS will be interesting to identify in the future.

TORC1 is the major controller of cellular metabolism that links environmental cues and signals for cellular growth and homeostasis. TORC1 upregulates anabolic processes such as de novo synthesis of proteins, nucleotides, lipids, and downregulates catabolic processes such as inhibition of autophagy 25,27. We find that dysfunctional mitochondria lead to TORC1 activation both in yeast and in human cells with accompanying onset of accelerated cellular aging. Our results are in line with a recent report about anabolic pathways enhancement and suppression of catabolic processes in cells with defective mitochondria 61. On the other hand, TORC1 activity decreases under enhanced mitochondrial function environment. The cause of the mitochondrial dysfunction (whether due to deletion of a critical gene or pharmacological intervention) is irrelevant in this context. Nevertheless, despite the mitochondrial insufficiency background, inhibition of TORC1 (e.g., with rapamycin) can partially rescue the shortened cellular lifespan (Figures 6D-6H; and S6F).

Our work sheds light onto the questions (i) how mitochondrial dysfunction is linked to accelerated cellular aging and (ii) how TORC1 inhibition suppresses the mitochondrial dysfunction and prevents shortening lifespan. Apparently, mitochondrial dysfunction aberrantly signals to increase the TORC1 activity that leads to accelerated aging in cells. Remarkably, TORC1 inhibition can often suppress the accelerated cellular aging associated with impaired mitochondrial function. Yet, we found that growth of mutant strains with dysfunctional mitochondria can be resistant despite TORC1 inhibition by rapamycin with concentrations effective in wild type cells possibly due to the dramatic increase of TORC1 activity as in YBR238C-OE, rmd9Δ, pet100Δ, cox6Δ and hap4Δ cells (Figures S6G and S6H).

Our interpretation of the experimental results is supported by two recently published studies: (1) TORC1 activation was observed after mitochondrial ETC dysfunction in an induced pluripotent cell model 62. (2) The inhibition of TORC1 delays the progression of brain pathology of mice with ETC complex I NDUFS4-subunit knockout 63.

We think that it will be insightful to explore TORC1 activity under various conditions with compromised mitochondrial function including mitochondrial fission and fusion dynamics which is reported to affect in cell viability 64,65. Altogether, our findings uncover the central role of the feedback loop between mitochondrial function and TORC1 signaling (TORC1-MItochondria-TORC1 (TOMITO) signaling process). Whereas the effector from TORC1 to mitochondria involves YBR238C, the other direction appears executed via metabolite sensing (e.g., α-keto-glutarate, glutamine, etc.) 66.

Methods

Data acquisition

The gene lists that modulate cellular lifespan in aging model organism yeast Saccharomyces cerevisiae were extracted from database SGD 21 and GenAge 22 (as of 8th November 2022). The actual gene lists are available in ‘Additional File 1’.

Yeast strains, growth media and cell culture

The S. cerevisiae auxotrophic BY4743 (Euroscarf) and prototrophic CEN.PK113-7D 67 strains were used in this study. Deletion strains for BY4743 obtained from yeast homozygous diploid collection. Deletion strains in CEN.PK background was generated using standard PCR-based method 68. Yeast strains were revived from frozen glycerol stock on YPD agar (1% Bacto yeast extract, 2% Bacto peptone, 2% glucose and 2.5% Bacto agar) medium for 2-3 days at 30°C. For all experiments, CEN.PK113-7D strains were grown in synthetic defined (SD) medium contain 6.7 g/L yeast nitrogen base with ammonium sulfate without amino acids and 2% glucose. SD medium supplemented with histidine (40 mg/L), leucine (160 mg/L), and uracil (40 mg/L) for auxotrophic BY4743 strains.

Human cell lines, growth media and cell culture

Human embryonic kidney cell line (HEK293, ATCC) was cultured in high-glucose DMEM supplemented with 10% FBS and 1% Penicillin Streptomycin Solution. All cells cultured in a humidified incubator with 5% CO2 at 37°C.

Chemical treatment to cell culture

Stock solution of rapamycin and antimycin A was prepared in dimethyl sulfoxide (DMSO). The final concentration of DMSO did not exceed 1% in yeast and 0.01% in human cell lines experiments.

Yeast aging assay

For the chronological lifespan (CLS) experiments, prototrophic CEN.PK113-7D strains were grown in synthetic defined (SD) medium contain 6.7 g/L yeast nitrogen base with ammonium sulfate without amino acids and 2% glucose. SD medium supplemented with histidine (40 mg/L), leucine (160 mg/L), and uracil (40 mg/L) for auxotrophic BY4743 strains. Chronological aging was assessed by determining the lifespan of yeast as described previously with slight modifications 15,69. Yeast culture grown in overnight at 30°C with 220 rpm shaking in glass flask was diluted to starting optical density at 600 nm (OD600) ∼ 0.2 in fresh medium to initiate the CLS experiment. CLS was performed by outgrowth method utilizing three different approaches: (i) Cells grown and aged in 96-well plates with a total 200 µL culture in SD medium at 30°C. At various age time points yeast stationary culture (2μL) were transferred to a second 96-well plate containing 200 μL YPD medium and incubated for 24 hours at 30°C without shaking. Outgrowth OD600 for each age point was measured by the microplate reader; (ii) Cells grown and aged in flask with a total culture volume more than 5 ml SD medium and incubated for 24 hours at 30°C with 220 rpm shaking. At different age time points yeast stationary culture washed and normalized to OD600 of 1.0 with YPD medium. Further, normalized yeast cells were serial 10-fold diluted with YPD medium in 96-well plates. 3 μL of diluted culture were spotted onto the YPD agar plate and incubated for 48 hours at 30°C. The outgrowth of aged cells on the YPD agar plate was photographed using the GelDoc imaging system; (iii) The above discussed serial 10-fold diluted yeast stationary culture with YPD medium in 96-well plates incubated for 24 hours at 30°C without shaking. Outgrowth OD600 for serial diluted aged cells was measured by the microplate reader.

RNA extraction

Yeast cells were first mechanically lysed using the manufacturer’s disruption protocol. Total RNA from yeast cells was extracted using Qiagen RNeasy mini kit. ND-1000 UV-visible light spectrophotometer (Nanodrop Technologies) and Bioanalyzer 2100 with the RNA 6000 Nano Lab Chip kit (Agilent) was used to assess the concentration and integrity of RNA.

RNA sequencing and bioinformatics analysis

RNA sequencing (RNA-Seq) was conducted using NovaSeq PE150. Raw Fastq files were then passed into Fastp v0.23.2 for adapter trimming and low-quality reads removal. Both single end and paired end reads were processed with default parameters with --detect adapter for pe. Raw reads that passed the quality check were then aligned using HiSat 2 v2.2.1 with the index built from Saccharomyces cerevisiae R-64-1-1 top-level DNA fasta file obtained from Ensembl for sequencing data with S288C strain background. Library information for all experiments were first checked with RSeQC infer_experiment.py v4.0.0 before the addition of the respective parameters for alignment and counts. The resulting SAM files were then converted and sorted to BAM files using Samtools v1.13. The BAM files were then used to generate feature counts using HTSeq v1.99.2. HTSeq counts from each experiment were then used for downstream Differentially Expressed Genes (DEG) analysis. Counts generated from HTSeq were then used for differential gene expression analysis using EdgeR v3.34.1 quasi likelihood F-test and DESeq2 v1.36.0. Principal Component Analysis (PCA) was conducted via the use of Transcript Per Million normalized counts. Samples that are separated by batches in the PCA were corrected using ComBat-Seq v3.44.0 before PCA was conducted after normalization of the corrected counts. Functional enrichment analysis was performed by metascape tool 28.

qRT-PCR analysis

qRT-PCR (Quantitative Reverse Transcription-Polymerase Chain Reaction) experiments were performed as described previously using QuantiTect Reverse Transcription Kit (Qiagen) and SYBR Fast Universal qPCR Kit (Kapa Biosystems) 33. The abundance of each gene was determined relative to the house-keeping transcript ACT1.

ATP analysis

ATP analysis was performed as described previously 70. Yeast cells were mixed with the final concentration of 5% trichloroacetic acid (TCA) and then kept on ice for at least 5 min. Cells were washed and resuspended in 10% TCA and lysis was performed with glass beads in a bead beater to extract the ATP. ATP extraction from human HEK293 cells was performed using Triton X-100 lysis buffer. The ATP level was quantified by PhosphoWorks™ Luminometric ATP Assay Kit (AAT Bioquest) and normalized by protein content measured Bio-Rad protein assay kit.

Mitochondrial DNA copy number analysis

Determination of mitochondrial DNA (mtDNA) copy number was analysed by real time qPCR (Quantitative-Polymerase Chain Reaction). DNA was extrcated using Quick-DNA Midiprep Plus Kit (Zymo Research). qPCR was performed in a final volume of 20 μl containing 20 ng of total DNA using SYBR Fast Universal qPCR Kit (Kapa Biosystems) and analysed using the Quant Studio 6 Flex system (Applied Biosystems). The real-time qPCR conditions were one hold at (95 °C, 180 s), followed by 40 cycles of (95 °C, 1 s) and (60 °C, 20 s) steps. A melting-curve analysis was included in the cycle after amplification to verify PCR specificity and the absence of primer dimers. Relative mtDNA content was determined by qPCR of mitochondrial genes (ATP6 and COX3) and normalized with nuclear specific gene ACT1.

ROS measurement

Cells were washed and resuspended in 1× phosphate buffer saline (PBS, pH 7.4). After that cells were incubated with 40 μM H2DCFDA (Molecular probe) for 30 min at 30°C 71. Cells were then washed with PBS and ROS level was measured by fluorescence reading (excitation at 485 nm, emission at 524 nm) by the microplate reader. The fluorescence intensity was normalized with OD600.

Transcription factors analysis

Transcription factors (TFs) enrichment analysis was performed using YEASTRACT 72,73. The significant p-value <0.05 considered for regulatory network analysis based on DNA-binding plus expression evidence. The TFs regulatory networks were visualized with a force-directed layout.

Fermentative and respiratory growth assay

Yeast cells grown in YPD medium were washed and normalized to OD600 of 1.0 with water. Further, normalized yeast cells were serial 10-fold diluted with water. 3 μL of diluted culture were spotted onto the agar medium YPD (2% glucose) and YPG (3% glycerol) and incubated for 48 hours at 30°C. The cell growth on the agar plate was photographed using the GelDoc imaging system.

Growth sensitivity assay

Effect of chemical compounds on cell growth was carried out in 96-well plates. At an OD600 of ∼ 0.2 in SD medium 200 µL yeast cells was transferred into the 96-well plate containing serially double-diluted concentrations of compounds. Cells were incubated at 30°C and the growth was measured at OD600 by the microplate reader.

Quantification and statistical analysis

Data analysis of all the experimental results such as mean value, standard deviations, significance, and graphing were performed using GraphPad Prism v.9.3.1 software. The comparison of obtained results were statistically performed using the Student’s t-tests, Ordinary One-way ANOVA and Two-way ANOVA followed by multiples comparison tests. In all the graph plots, P values are shown as *P < 0.05, **P < 0.01, ***P < 0.001, and ****P < 0.0001 were considered significant. ns: non-significant.

Supporting information